### 2020-04-08 README.Rmd

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rmimixpackage

<!-- badges: start -->
<!-- badges: end -->

# introduction 
The goal of rmimixpackage is to 
implement an R version of the mimix Stata package.This Imputes missing numerical outcomes 
for a longitudinal trial with protocol deviation under distinct treatment arm-based assumptions for the unobserved data, following the general algorithm of Carpenter, Roger, and Kenward (2013).

# imputation methods
The imputation methods are
    jump to reference (j2r), 
    copy increments in reference (cir)
    copy reference (cr).
 (when specifying these in the  methods argument then the reference treatment arm must be      specified in the refer argument).

     missing at random (mar),
    last mean carried forward (lmcf),

   As well these options a
   further option can be made specifying the causal model (see White, Joseph and Best, 2019)

  Note that  
    Individual-specific imputation methods (and their reference groups) can also be specified when there exists columns
    in the input data-set. However this is mutually exclusive of using overall methods and both should not simultaneously be specified.
    See examples below

## Installation

You can install the released version of rmimixpackage from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("rmimixpackage")
```

## Examples

This is a basic example which shows you how to solve a common problem:

```{r example}
library(rmimixpackage)
## basic example code
 #recode treatment groups to 1,2..
asthma$treat<-ifelse(asthma$treat==2,1,ifelse(asthma$treat==3,2,99))

# method =J2R and refer =1
impdataset<-(Runmimix("asthma",c("base"),"fev","treat","id","time",10,1,"J2R",101,"jeffreys",1000,NULL,NULL) )

#anti-depressant data with methdovar & refernecevar
antidepressant<-read.table("http://raw.githubusercontent.com/UCL/mimix/master/data/SASantidepmethodvar.csv",header=TRUE,sep=",",fileEncoding = "UTF-8-BOM")

#recode treatment variable to numeric
antidepressant$TREATMENT.NAME<-ifelse(antidepressant$TREATMENT.NAME=="PLACEBO",1,ifelse(antidepressant$TREATMENT.NAME=="DRUG",2,99))
#  run individual-specific,  methodvar, 
impdataset <- Runmimix("antidepressant",c("basval"),"HAMD17.TOTAL","TREATMENT.NAME","PATIENT.NUMBER",
                       "VISIT.NUMBER",10,NULL,NULL,101,c("jeffreys"),1000,NULL,c("methodvar","referencevar"))

# run overall methd = J2R, refer =2 treatment group , ridge
impdataset <- Runmimix("antidepressant",c("basval"),"HAMD17.TOTAL","TREATMENT.NAME","PATIENT.NUMBER",
                             "VISIT.NUMBER",1000,2,"J2R",101,c("ridge"),1000,NULL,NULL)
```

What is special about using `README.Rmd` instead of just `README.md`? You can include R chunks like so:

```{r cars}
summary(cars)
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub!

### 2021-02-22 paper/paperRmimix.tex
First incomplete draft of SJ paper

% !TeX root = RJwrapper.tex

% article type: add-on package. 
%``short introductions to contributed R packages that are already available on
%CRAN or Bioconductor, and going beyond package vignettes in aiming to provide broader
%context and to attract a wider readership than package users. Authors need to make a strong
%case (in a motivating letter accompanying a submission) for such introductions, based for
%example on novelty in implementation and use of R, or the introduction of new data structures
%representing general architectures that invite re-use. Authors of narrower package-introduction
%articles may wish to consider alternatives such as The Journal of Open Source Software (http:
%//joss.theoj.org/) or in the life sciences, the F1000 Bioconductor (https://f1000research.
%com/channels/bioconductor) or Rpackage (https://f1000research.com/channels/rpackage)
%channels.''

% to do: 
%Create a minimal ‘filename.bib’ BibTeX file with only the entries you need, without unnecessary
%fields such as abstracts, and add \bibliography{filename} at the end of ‘filename.tex’, where
%filename is the same file name created above.

% to do at end:
% produce shortened .bib & change bibliography cmd
%Run tools::texi2pdf on ‘RJwrapper.tex’ to produce ‘RJwrapper.pdf’, using clean=FALSE to help
%in debugging if necessary.
%Iterate until ‘RJwrapper.pdf’ looks right.

%\newcommand {\E}      [1] {{\mbox{E}}\left[#1\right]}
\newcommand {\varb}   [1] {{\rm var}\left(#1\right)}
\newcommand {\cov}    [2] {{\rm cov}\left(#1,#2\right)}
\newcommand {\cind}      {\mbox{$\perp\hspace{-0.5em}\perp$}}
\newcommand {\logit}  [1] {{\rm logit}\ #1}
\newcommand {\twovec}[2] {\left( \begin{array}{c} #1 \\ #2 \end{array} \right)}
\newcommand {\N} [2] {N\left(#1,#2\right)}

\newcommand {\pret}{{\le{t}}}
\newcommand {\postt}{{>t}}
\newcommand{\bX}{\boldsymbol{X}}
\newcommand{\bY}{\boldsymbol{Y}}
\newcommand {\tmax}{T} % {t_{max}}
\newcommand{\KK}{k_1}

\title{mimix: Reference-based imputation of missing data}
\author{by Ian R White, Kevin McGrath, Matteo Quartagno, Suzie M Cro and James Carpenter}

\maketitle

\abstract{
Reference-based imputation is a multiple imputation technique which imputes quantitative outcome data that are missing after participant discontinuation of allocated treatment in a randomised trial.
We present and describe an R package, mimix, for performing reference-based imputation, including a causal model variant, and we compare its implementation with that in SAS and Stata.
}

\section{Introduction}

Missing data are a challenge for many analyses. This article tackles the specific issue of a randomised trial with a repeatedly measured quantitative outcome, where participants who discontinue their randomised treatment are not followed up thereafter and hence have missing outcome data. 
We assume that the aim is to estimate the effect of treatment on the actual outcomes of the participants, whether observed or not: this has been termed a ``de facto'' estimand \citep{Carpenter++13} or a ``treatment policy'' estimand \citep{ICHE9R1}. 
In this setting, an analysis under the commonly used missing at random (MAR) assumption would assume that the missing (post-treatment) outcomes are comparable (conditional on observed data) with the observed (on-treatment) outcomes, and would therefore estimate the effect of treatment on the outcomes of the participants if treatment was never discontinued: this has been termed a ``de jure'' estimand \citep{Carpenter++13} or a ``hypothetical'' estimand \citep{ICHE9R1}.

% methods
\citet{Carpenter++13} proposed that missing data after discontinuation of allocated treatment could be imputed by assuming that post-discontinuation outcomes behave, in some sense, like the outcomes in a reference group. For example, if participants who have discontinued their allocated treatment are likely to be receiving similar treatment to participants who were allocated to control treatment, then the control group would be the reference group.
\citet{Carpenter++13} proposed imputing the missing data from a joint multivariate Normal (MVN) distribution for the complete (observed and unobserved) data, and proposed five ways to construct this joint distribution: ``jump to reference'' (J2R), ``copy reference'' (CR), ``copy increments in reference'' (CIR), ``last mean carried forward'' (LMCF) and ``missing at random'' (MAR). All approaches start by fitting a MVN distribution to the data from each arm. As an example, the J2R joint distribution for a participant in a specific arm takes the means for that arm up to the point of treatment discontinuation and the means for the reference arm afterwards.

The implicit assumptions behind this approach were explored by \citet{ian:RBIcausal}, who proposed a causal model in which the treatment effect after discontinuation is a specified multiple of the treatment effect at the point of discontinuation. They showed that J2R, CR and CIR are the special cases of this causal model in which the treatment effect disappears, decays or is maintained after treatment discontinuation, while LMCF and MAR are not special cases of the causal model.

% software 
The RBI methods of \citet{Carpenter++13} were implemented in SAS in what have become known as ``the five macros'' and are available on the web page (on www.missingdata.org.uk) of the DIA working group for missing data. These macros also provide for ``delta-adjustment'' in which imputed values are modified by a user-specified amount \citep{Ratitch++13}: this is useful is performing sensitivity analysis, since the RBI methods make a number of untestable assumptions \citep{ian:ZAMSTAR}.
The methods were then implemented in Stata by \citet{Cro++16}.

% Aims of paper
We have implemented the RBI methods in a new R package, which includes the a full implementation of the causal model. 
The causal model was previously specified for a two-arm trial and has been extended here for a multi-arm trial.
The aim of this paper is to describe our implementation of mimix in R, to illustrate its use, and to describe its novel features by comparison with the implementations in SAS and Stata.

% \section{Another section}
%
%This section may contain a figure such as Figure~\ref{figure:rlogo}.
%
%\begin{figure}[htbp]
%  \centering
%  \includegraphics{Rlogo-5}
%  \caption{The logo of R.}
%  \label{figure:rlogo}
%\end{figure}

%\section{Another section}
%
%There will likely be several sections, perhaps including code snippets, such as:
%
%\begin{example}
%  x <- 1:10
%  result <- myFunction(x)
%\end{example}

%This file is only a basic article template. For full details of \emph{The R Journal} style and information on how to prepare your article for submission, see the \href{https://journal.r-project.org/share/author-guide.pdf}{Instructions for Authors}.

\section{Reference-based imputation}

\subsection{Setting} [nicked from \citet{ian:RBIcausal}]
We assume that quantitative outcome measurements are scheduled at baseline and at $\tmax$ occasions after randomisation. 
Let $Z$ be the random variable for the participant's randomised treatment arm, let $Z=z$ denote an arm in which we want to impute missing values, and $Z=r$ denote the reference arm.
Let $Y_t$ be the random variable for the participant's outcome at visit $t=0,...,\tmax$. 
It is convenient to define 
$\boldsymbol{Y}_{\pret}=(Y_0,\ldots,Y_t)$, the vector of all outcomes up to  and including visit $t$; 
$\boldsymbol{Y}_{\postt}=(Y_{t+1},\ldots,Y_\tmax)$, the vector of all outcomes after visit $t$; 
and
$\boldsymbol{Y}=(Y_0,\ldots,Y_\tmax)$, the vector of all outcomes.
Let $D$ be the random variable for the participant's last visit prior to discontinuing treatment, so $D=0,...,\tmax$.
$Y_t$ is observable for all $t$ but only observed for $t \le D$, because we assume no off-treatment data. 
We aim to impute the unobserved values of $Y_t$ for $t>D$: we stress that these are the outcomes that existed but were unobserved, not the outcomes that would have existed if treatment had been continued.

\subsection{Reference-based imputation}

\citet{Carpenter++13}  proposed a generic MI algorithm for this setting:
\begin{enumerate}

\item For each treatment arm $z$, fit a multivariate normal model to all observed data, using a
    Bayesian approach with an improper prior and assuming MAR. The model has unstructured
    mean $\boldsymbol{\mu}_z$ and unstructured variance-covariance matrix $\boldsymbol{\Sigma}_z$.

\item For each treatment arm $z$, draw a mean vector $\boldsymbol{\mu}_z^*$ and variance-covariance matrix from the
    posterior distribution $\boldsymbol{\Sigma}_z^*$.

\item For each treatment arm  $z$ and each possible treatment discontinuation visit $t$, use the drawn values to build a hypothetical  joint  distribution of 
the outcomes $\boldsymbol{Y}_{\pret}$ up to time $t$ 
and 
the outcomes $\boldsymbol{Y}_{\postt}$ after time  $t$, 
using one of the methods described below.
Thus a MVN distribution is built for $\boldsymbol{Y}|Z=z,D=t$.
Five methods are mainly distinguished by their choice of mean: 
\begin{enumerate}
\item Jump to reference (J2R):
mean = $\left( \boldsymbol{\mu}_{z,\pret}^*, \boldsymbol{\mu}_{r,\postt}^* \right)$.

\item Copy reference (CR):
mean = $\boldsymbol{\mu}_r^*$.

\item Copy increments in reference (CIR):
mean = $\left( \boldsymbol{\mu}_{z,\pret}^*,  \{\mu_{z,t}^* - \mu_{r,t}^* \} \boldsymbol{e}_{\tmax-t} + \boldsymbol{\mu}_{r,\postt}^*   \right)$
where $\boldsymbol{e}_p$ is a row vector $(1,\ldots,1)$ of length $p$.

\item Missing at random (MAR): mean = $\boldsymbol{\mu}_z^*$.

\item Last mean carried forward (LMCF):
mean = $\left( \boldsymbol{\mu}_{z,\pret}^*, \mu_{z,t}^*\boldsymbol{e}_{\tmax-t} \right)$.
\end{enumerate}

CRK proposed corresponding variance matrices. Following \cite{ian:RBIcausal}, we describe only the regression coefficient matrix and conditional variance matrix of the potential outcomes after visit  $t$ given those before visit $t$. 
CRK set these to be taken from arm $z$ for MAR and LMCF, 
and from arm $r$ for J2R, CIR and CR.

[KEVIN: An approach that we call \emph{RBI alternative} instead uses
$\boldsymbol{\beta}_t(\tmax)$ and $\boldsymbol{\Omega}_t(\tmax)$ for all RBI methods: did we code this???]

\item For each treatment arm $z$ and each observed treatment discontinuation visit $t$, construct the imputation distribution of  $\boldsymbol{Y}_{\postt}$ given  $\boldsymbol{Y}_{\pret}$. 
Sample $\boldsymbol{Y}_{\postt}$ from this   conditional distribution, to create a “completed” data set.

\item Repeat steps 2-4 $m$ times, resulting in $m$ imputed data sets.

\item Analyse each imputed data set and combine the results using Rubin's rules \citep{Rubin87}.

\end{enumerate}

\subsection{Causal model}

The causal model has previously been stated for two arms only. Here we extend its statement to the multi-arm case. 
% Novelties: K depends on z as well as t in general
We define the potential outcome $Y_t(z,s)$ at visit $t$ as the outcome that would have been observable if, possibly contrary to fact, a participant received active treatment $z$ for $s$ periods only.
In particular, $Y_t(z,0)$ is the potential outcome if never treated: it is the same for all $z$ and is written $Y_t(0)$. Similarly $Y_t(z,\tmax)$ is the potential outcome if always treated with $z$.
We define 
$\boldsymbol{Y}_{\pret}(z,s)$, $\boldsymbol{Y}_{\postt}(z,s)$ and $\boldsymbol{Y}(z,s)$ as before.
We let 
$\mu_t(z,s) = \E[Y_t(z,s)]$, the mean of the potential outcome at visit $t$ if active treatment $z$ were received for $s$ periods only. Similarly we define 
$\boldsymbol{\mu}_\pret(z,s)$, $\boldsymbol{\mu}_\postt(z,s)$ and $\boldsymbol{\mu}(z,s)$.

[Omit?:] 
The variance-covariance matrix  of the potential  outcomes is $\boldsymbol{\Sigma}(s)=\varb{\boldsymbol{Y}(s)}$ with submatrices 
$\boldsymbol{\Sigma}_{\pret\pret}(s)$,
$\boldsymbol{\Sigma}_{\postt\postt}(s)$ and
$\boldsymbol{\Sigma}_{\postt\pret}(s)$.
We define 
the matrix of regression coefficients of  potential outcomes after visit  $t$ on those up to visit  $t$
as $\boldsymbol{\beta}_t(s) = \boldsymbol{\Sigma}_{\postt\pret}(s) \boldsymbol{\Sigma}_{\pret\pret}(s)^{-1}$, and 
the residual variance of the potential outcomes after visit  $t$ given those up to visit  $t$ as
$\boldsymbol{\Omega}_t(s) = \boldsymbol{\Sigma}_{\postt\pret}(s)
\boldsymbol{\Sigma}_{\pret\pret}(s)^{-1}
\boldsymbol{\Sigma}_{\postt\pret}(s)^T$.

The key model assumption describes how the maintained effect of treatment after
discontinuation relates to the effect of treatment before discontinuation:
\begin{equation}\label{eq:del}
\boldsymbol{\mu}_{\postt}(z,t)-\boldsymbol{\mu}_{\postt}(0) =
\boldsymbol{K}_{z,t} \left\{\boldsymbol{\mu}_{\pret}(z,t)-\boldsymbol{\mu}_{\pret}(0)\right\}
\end{equation}
where $\boldsymbol{K}_{z,t}$ is a $(\tmax - t) \times (t+1)$ matrix of sensitivity parameters: it is not
identified by the data and must be specified by the user.
In practice we make a choice of $\boldsymbol{K}_{z,t}$ determined by just two parameters $(k_0,k_1)$ giving the simplified model for each $u>t$:
\begin{eqnarray}
\boldsymbol{\mu}_u(z,t) - \boldsymbol{\mu}_u(0)  & =& k_0 k_1^{v_u-v_t} \left\{ \boldsymbol{\mu}_t(z,t) - \boldsymbol{\mu}_t(0) \right\} \label{eq:mte3}
\end{eqnarray}
where  $v_t$ is the time (on a suitable scale) of visit $t$.

[KEVIN - can we let $(k_0, k_1)$ depend on arm?]

Estimation involves other assumptions which make explicit the ideas of \citet{Carpenter++13}: 
that randomisation is independent of all potential outcomes; 
that step 1 of the RBI algorithm in the active arms estimates the distribution of $\boldsymbol{Y}(z,s)$,
and in the reference arm estimates the distribution of $\boldsymbol{Y}(0)$; 
that the conditional distributions follow linear regressions (which is true if the joint distribution is MVN); 
and that treatment discontinuation is unaffected by future potential partly-treated outcomes.

These give the mean of the imputation model:
\begin{eqnarray}\label{eq:causal}
&&E[\boldsymbol{Y}_{\postt}(z,t) | Z=z, \boldsymbol{Y}_{\pret}, D=t]
\nonumber\\
&&=
\boldsymbol{\beta}_{t}(z,t) \left\{\boldsymbol{Y}_{\pret} - \boldsymbol{\mu}_{\pret}(z,t)\right\}  +
\boldsymbol{K}_{z,t} \left\{\boldsymbol{\mu}_{\pret}(z,t)-\boldsymbol{\mu}_{\pret}(0)\right\} +
\boldsymbol{\mu}_{\postt}(0).
\end{eqnarray}

[Again we can handle variances... how much to say?]

\subsection{Delta adjustment}

We follow the approach of the five macros. Any imputed value $Y^*_{z,t}$ is replaced by $Y^*_{z,t} + \sum_{u=t+1}^{u=\tmax} a_u b_{u-t}$ 
where $a_u$ is a user-specified shift for time $u$ and $b_w$ is a user-specified scaling multiplier that controls how the user-defined shift for time $u$ is applied to all times $\ge u$. For example, if $b=(1,0.7, 0.5,...)$ then a participant who discontinues treatment just after visit $t$ has 
their imputed value at time $t+1$ incremented by $a_{t+1}$, 
their imputed value at time $t+2$ incremented by $0.7 a_{t+1} + a_{t+2}$, 
their imputed value at time $t+3$ incremented by $0.5 a_{t+1} +0.7  a_{t+2} + a_{t+3}$, and so on. 
Delta adjustment can apply either with RBI or the causal method and is always applied after imputation.

\subsection{Interim missing values and covariates}

[Kevin - to discuss]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Package}

KEVIN TO DRAFT - INITIALLY TAKE FROM HELP.

\subsection{Arguments}
\subsection{Outputs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Examples}

KEVIN TO DRAFT. Which data to use?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Comparisons with other packages}

KEVIN TO DRAFT -- THESE ARE JUST MY NOTES. Would a table be suitable?

\begin{enumerate}
\item flavours of LMCF in 5 macros
\item handling participants with no observed outcomes
\item how baseline covariates are modelled: SAS, can interact with time; Stata, R, can't
\item handling of interim missing values
\item anything to say about algorithm differences? prior choice?
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Limitations and discussion}

\begin{itemize}
\item flexible package
\item doesn't cover other outcome types; e.g. methods have been proposed for recurrent events \citep{Keene++14}
\item trials should be designed consistently with their estimand, so if the treatment-policy estimand is of interest then outcomes should be collected after treatment discontinuation. Analysis options for this setting are still in development: options include including treatment discontinuation time in the model and imputing under MAR [ref Tom at GSK]; using RBI or causal model and reserving the post-discontinuation data for model checking; or using RBI or causal model and using the post-discontinuation data to estimate model parameters such as $K_{z,t}$. 
\end{itemize}

%\bibliography{RJreferences}
%\bibliographystyle{sj}
\bibliography{c:/latex/library}

\address{Ian R White\\
MRC Clinical Trials Unit at UCL\\
90 High Holborn, 2nd Floor, London WC1V 6LJ\\
UK\\
ORCiD: 0000-0002-6718-7661\\
  \email{ian.white@ucl.ac.uk}}

\address{Kevin McGrath\\
MRC Clinical Trials Unit at UCL\\
90 High Holborn, 2nd Floor, London WC1V 6LJ\\
UK\\
  (ORCiD if desired)\\
  \email{kevin.mcgrath@ucl.ac.uk}}

\address{Matteo Quartagno\\
MRC Clinical Trials Unit at UCL\\
90 High Holborn, 2nd Floor, London WC1V 6LJ\\
UK\\
  (ORCiD if desired)\\
  \email{m.quartagno@ucl.ac.uk}}

\address{Suzie M Cro\\
Imperial Clinical Trials Unit \\
Imperial College London, 1st Floor, Stadium House
London, W12 7RH\\
  UK\\
  (ORCiD if desired)\\
  \email{s.cro@imperial.ac.uk}}

\address{James Carpenter\\
MRC Clinical Trials Unit at UCL\\
90 High Holborn, 2nd Floor, London WC1V 6LJ\\
UK\\
  (ORCiD if desired)\\
  \email{j.carpenter@ucl.ac.uk}}

### 2021-02-23 R/Runmimix.R
manual pdf

@@ -1,10 +1,11 @@
-#' @title mimix
+#' @title Main function for performing reference-based multiple imputation of longitudinal data  
 #' @description main wrapper for running mimix
 #' @details This is based on Suzie Cro's Stata program
 #' @details sets up a summary table based on missing data pattern- mg  mimix_group                                                                   
 #' @details reflects the pattern and treatment group configuration of the raw data
 #' @details then acts as a looping mechanism, norm2 is used as MCMC multivariate normal 
 #' @export mimix
+#' @import mice
 #' @param data  Dataset in wide format
 #' @param covar Covariates - may include the baseline value of depvar. Must be complete (no missing values).
 #' @param depvar Dependent (outcome) variable
@@ -28,9 +29,12 @@
 #' @return impdataset the M imputed data-sets appended to the "missing values" data-set in wide format
 #' @examples
 #' \dontrun{
-#' mimix("asthma",c("base"),"fev","treat","id","time",5,1,"J2R",,,,,,,,,1,0.5,)
-#' fit<-with(data= as.mids(impdataCausal), expr = lm(fev.12~treat+base))
+#' mimixout<-mimix("asthma",c("base"),"fev","treat","id","time",5,1,"J2R",,,,,,,,,1,0.5,)
+#' library(mice)
+#' fit<-with(data= as.mids(mimixout),expr = lm(fev.12~treat+base))
 #' summary(pool(fit))
+#' mimix("acupuncture",c("head_base"),"head","treat","id","time",1000,1,"CIR",54321,
+#'     "jeffreys",1000,NULL,NULL,NULL,NULL,NULL,K0=1,K1=1,mle=0  )
 #' }
 
 mimix<- function(data,covar=NULL,depvar,treatvar,idvar,timevar,M=1,reference=NULL,method=NULL,seed=101,prior="jeffreys",burnin=1000,bbetween=NULL,methodvar=NULL,referencevar=NULL,delta=NULL,dlag=NULL,K0=1,K1=1,mle=FALSE) {
@@ -530,7 +534,7 @@ mimix<- function(data,covar=NULL,depvar,treatvar,idvar,timevar,M=1,reference=NUL
                   # in case more than 1 interim in patt group
                     for (it_interim in 1:mg[i,"cases"]) {
                     
[…]

### 2021-02-23 R/data.R

@@ -57,20 +57,21 @@
 #'  }
 "acupuncture"
 
-#' impdataCausal.
-#' 
-#'  A dataset containing imputed values obtained from mimix on asthma data 
-#'  
-#'   @format A data frame with 1098 rows and 9 columns
-#'   \describe{
-#'   \item{.imp}{imputation number}
-#'   \item{base}{covariate}
-#'   \item{fev.2}{response at time 2}
-#'   \item{fev.4}{response at time 4}
-#'   \item{fev.8}{response at time 8}
-#'   \item{fev.12}{response at time 12}
-#'   \item{patt}{missing data pattern}
-#'   \item{treat}{treatment group}
-#'   \item{.id}{case id}
-#'   }
-"impdataCausal"   
+# eample output file 
+# impdataCausal.
+# 
+#  A dataset containing imputed values obtained from mimix on asthma data 
+#  
+#   @format A data frame with 1098 rows and 9 columns
+#   \describe{
+#   \item{.imp}{imputation number}
+#   \item{base}{covariate}
+#   \item{fev.2}{response at time 2}
+#   \item{fev.4}{response at time 4}
+#   \item{fev.8}{response at time 8}
+#   \item{fev.12}{response at time 12}
+#   \item{patt}{missing data pattern}
+#   \item{treat}{treatment group}
+#   \item{.id}{case id}
+#  }
+#"mpdataCausal"   

### 2021-02-23 R/proprocess.R

@@ -1,6 +1,6 @@
 
 
-#' @title preprodata
+#' @title pre-process long longitudinal data into wide format 
 #' @description process data into wide format for group method
 #' @details checks method finds missingness pattern
 #' @param data  data in long format
@@ -201,7 +201,7 @@ preprodata<- function(data,covar,depvar,treatvar,idvar,timevar,M,reference,metho
 
 }
 
-#' @title preproIndivdata
+#' @title pre-process long longitudinal data into wide format for individual-specific  
 #' @description process data into wide format for individual-specified method
 #' @details checks methodvar finds missingness pattern
 #' @param data  data in long format
@@ -414,7 +414,7 @@ preproIndivdata<- function(data,covar,depvar,treatvar,idvar,timevar,M,reference=
 
 
 
-#' @title ifmethodindiv
+#' @title performs imputation for individual-specific method
 #' @description alternative logic for individual method
 #' @details checks methodindiv not null
 #' @param methodvar  individual method col

### 2021-02-23 R/utilities.R

@@ -47,9 +47,11 @@
 #library(amelia)
 
 
-#' @title LMCF_loop
+#' @title process LMCF method
 #' @description process LMCF method
 #' @details This is based on Suzie Cro's Stata program
+#' @details when no observed data first mean is used in Stata 
+#' @details may be different here  
 #' @param c_mata_miss vector of col locaton of missing values , eg 5 6  
 #' @param mata_Means vector of means after mcmc draws eg 17 1 16.8 15.5 14.6 13.2 
 #' @return mata_means
@@ -66,7 +68,7 @@ LMCF_loop <- function(c_mata_miss,mata_Means)
   return(mata_means)
 }
 
-#' @title CIR_loop
+#' @title process CIR method
 #' @description process CIR method
 #' @details This is based on Suzie Cro's Stata program
 #' @param c_mata_miss vector of col locaton of missing values , eg 5 6  
@@ -124,7 +126,7 @@ CIR_loop <- function(c_mata_miss,mata_Means,MeansC)
 
 
 
-#' @title Causal_loop
+#' @title process Causal method
 #' @description process Causal method
 #' @details This is based on "White,Royes,Best" paper
 #' @param c_mata_miss vector of col locaton of missing values , eg 5 6  
@@ -268,7 +270,7 @@ Causal_loop<- function(c_mata_miss,mata_Means,MeansC,K0,K1)
 
 
 # dont include rhtis in the final package , jus for test purposes only!
-#' @title analyselist
+#' @title find descriptive stats on the  M imputed data set
 #' @description find descriptive stats on the  M imputed data set
 #' @details select on patient id and find their means etc
 #' @param id patient identifier
@@ -291,7 +293,7 @@ analyselist <-function(id,datlist,varlist) {
    t(round(pastecs::stat.desc(datano)[,varlist],8)[c(1,9,13,4,8,5),])
  }
 
-#' @title AddDelta
+#' @title ddd delta's to imputed values
 #' @description add delta's to imputed values
 #' @details Adding delta values after wthdrawal
 #'  Specifying delta and dlag allows imputations to differ sytematically from RBI methods. 

### 2021-02-23 proprocess.R
Delete proprocess.R

@@ -1,645 +1,0 @@
-
-
-#' @title preprodata
-#' @description process data into wide format for group method
-#' @details checks method finds missingness pattern
-#' @param data  data in long format
-#' @param covar covariates and base depvar complete
-#' @param depvar dependent variable
-#' @param treatvar treatment group
-#' @param idvar patient id
-#' @param timevar time variable for repeated visit
-#' @param M number imputations
-#' @param reference reference group
-#' @param method RBI method
-#' @return list of outputs
-
-
-#preprocess data for group method
-preprodata<- function(data,covar,depvar,treatvar,idvar,timevar,M,reference,method=NULL)  {
-  #extract relevant vars
-   
-  #to investigate interims invoke browser and edit(finaldat) 
-  #browser()
-  # if covar null (or 1st depvar complete) then ceate baseval covar
-
-  # extract relevant cols
-    # fevdata<-dplyr::select(get(data),idvar,depvar,timevar)
-  ##2311 make sure .id is id ! 
-  #browser()
-    fevdata<- get(data)[c(idvar,covar,depvar,timevar,treatvar)]
-  # extract covar cols 1 row per id to merge onto the wide data
-    uniqdat<-unique(get(data)[c(idvar,covar,treatvar)])
-    ntreatcol<- get(data)[c(treatvar)]
-    ntimecol<- get(data)[c(timevar)]
-
-
-
-  #replace tidyr function
-  #sts4<-tidyr::pivot_wider(fevdata,id_cols=c(idvar),names_from=timevar,names_prefix=depvar,values_from=depvar)
-
- # reshape from long to wide longitudinal data with as many depvars as time points
-  sts4<-stats::reshape(fevdata,v.names = depvar,timevar = timevar,idvar=idvar,direction="wide")
-
-  #assumes no NA for these vars should add in checking routines !!
-  #check how many covars used
-
-  #nocovar=covar[[1]]
-
- # print(paste0("covar=",covar))
-
- # print(utils::head(uniqdat))
-
-  finaldatOld<- merge(sts4,uniqdat,by=idvar)
-  # now no need to merge because covar ,treat already specified in fevdata
[…]

### 2021-02-25 R/Runmimix.R
comment browser

@@ -1908,14 +1908,14 @@ fillinterims<- function(impdata,interims,Mimp=M ) {
   impMarint_dt <- data.table::as.data.table(impdata)
   interims_dt <- data.table::as.data.table(interims)
   
-  #browser(text="setkey 1112")
+  #browser(text="2402")
   
   data.table::setkey(impMarint_dt,.id)
   data.table::setkey(interims_dt,V1)
   #merge
-  (impMarint_dt[interims_dt])
+  #(impMarint_dt[interims_dt])
   # convert to 0, 1 =missing
-  sapply(impMarint_dt[interims_dt],function(x) ifelse(is.na(x) ,1,0) )
+  # sapply(impMarint_dt[interims_dt],function(x) ifelse(is.na(x) ,1,0) )
   # exclude last non-response cols
   test10<-sapply(impMarint_dt[interims_dt],function(x) ifelse(is.na(x) ,1,0) )
   test10x<-test10[,c(1:(ncol(test10)-3))]

### 2021-03-02 R/Runmimix.R
editing after data.table match failed

@@ -37,6 +37,7 @@
 #'     "jeffreys",1000,NULL,NULL,NULL,NULL,NULL,K0=1,K1=1,mle=0  )
 #' }
 
+# v0.0.11
 mimix<- function(data,covar=NULL,depvar,treatvar,idvar,timevar,M=1,reference=NULL,method=NULL,seed=101,prior="jeffreys",burnin=1000,bbetween=NULL,methodvar=NULL,referencevar=NULL,delta=NULL,dlag=NULL,K0=1,K1=1,mle=FALSE) {
   # 6/11 try account for interims J2R MAR
   # if testinterims then want method to 1stly be MAR
@@ -358,10 +359,13 @@ mimix<- function(data,covar=NULL,depvar,treatvar,idvar,timevar,M=1,reference=NUL
   else {
     cat("\nImputing missing values using individual-specific method:\n\n")
   }
+  
+  #browser(text="0103")
   #initialise interim 
   interim<-0
   # construct structure to savde interim ids but this get reinitialised to many times!
   interim_id<- mata_Obs[c(mg[1,1]),"id"] 
+  
   #interim_pos <- c(0,0,0,0,0)
   
   #18/11 see if rawplusinterim works here!
@@ -542,13 +546,14 @@ mimix<- function(data,covar=NULL,depvar,treatvar,idvar,timevar,M=1,reference=NUL
                     # construct vector to save interims ids
                #11/11   interim_id <- rbind(interim_id, mata_Obs[c(mg[i,"cumcases"]),"id"] )  
                 
-                   
+                 # browser(text="0103") # check logic ,eg c_mata_miss 45  
                   
                 } else 
                 # note that if all missing then wont be interims and c_mata_nonmiss is integer(0) ie NUL
-          if (length(c_mata_nonmiss)!=0) {(c_mata_miss[b-1]+1 == c_mata_miss[b]) & ( c_mata_miss[b-1] < max(c_mata_nonmiss)) }   # need to check there is a non-missing to the right, eg 23 5 , 234 all interims!  
+          #if (length(c_mata_nonmiss)!=0) {(c_mata_miss[b-1]+1 == c_mata_miss[b]) & ( c_mata_miss[b-1] < max(c_mata_nonmiss)) }   # need to check there is a non-missing to the right, eg 23 5 , 234 all interims!  
[…]

### 2021-03-03 paper/paperRmimix.tex
Comments on manual and updates to paper based on manual

@@
 The RBI methods of \citet{Carpenter++13} were implemented in SAS in what have become known as ``the five macros'' and are available on the web page (on www.missingdata.org.uk) of the DIA working group for missing data.
-These macros also provide for ``delta-adjustment'' in which imputed values are modified by a user-specified amount \citep{Ratitch++13}: this is useful is performing sensitivity analysis, since the RBI methods make a number of untestable assumptions \citep{ian:ZAMSTAR}.
+These macros also provide for ``delta adjustment'' in which imputed values are modified by a user-specified amount \citep{Ratitch++13}: this is useful is performing sensitivity analysis, since the RBI methods make a number of untestable assumptions \citep{ian:ZAMSTAR}.
 The methods were then implemented in Stata by \citet{Cro++16}.
+A practical guide to their use is given by \citet{Cro++20}.
 
@@
 
-CRK proposed corresponding variance matrices.
-Following \cite{ian:RBIcausal}, we describe only the regression coefficient matrix and conditional variance matrix of the potential outcomes after visit  $t$ given those before visit $t$.
-CRK set these to be taken from arm $z$ for MAR and LMCF, 
-and from arm $r$ for J2R, CIR and CR.
-
-[KEVIN: An approach that we call \emph{RBI alternative} instead uses
-$\boldsymbol{\beta}_t(\tmax)$ and $\boldsymbol{\Omega}_t(\tmax)$ for all RBI methods: did we code this???]
+The variance matrices are constructed so that the regression coefficient matrix and conditional variance matrix of the potential outcomes after visit  $t$ given those before visit $t$ are taken from arm $z$ for MAR and LMCF, and from arm $r$ for J2R, CIR and CR
+\citet{Carpenter++13,ian:RBIcausal}.
+[NOTE we didn't code the \emph{RBI alternative} approach of \citet{ian:RBIcausal} that instead uses $\boldsymbol{\beta}_t(\tmax)$ and $\boldsymbol{\Omega}_t(\tmax)$ for all RBI methods.]
 
@@
 
-[KEVIN - can we let $(k_0, k_1)$ depend on arm?]
+[AT PRESENT $(k_0, k_1)$ ARE FIXED BUT IT MAY BE POSSIBLE TO LET THEM DEPEND ON ARM.
+THIS MAKES SENSE FOR A TRIAL WITH MORE THAN 2 ARMS.
+WOULD PROBABLY CODE THEM AS A VARIABLE IN DATA SET SO COULD LET THEM VARY BY OTHER FACTORS E.G.
+REASON?]
 
@@
 
-We follow the approach of the five macros. Any imputed value $Y^*_{z,t}$ is replaced by $Y^*_{z,t} + \sum_{u=t+1}^{u=\tmax} a_u b_{u-t}$ 
-where $a_u$ is a user-specified shift for time $u$ and $b_w$ is a user-specified scaling multiplier that controls how the user-defined shift for time $u$ is applied to all times $\ge u$.
-For example, if $b=(1,0.7, 0.5,...)$ then a participant who discontinues treatment just after visit $t$ has
-their imputed value at time $t+1$ incremented by $a_{t+1}$, 
-their imputed value at time $t+2$ incremented by $0.7 a_{t+1} + a_{t+2}$, 
-their imputed value at time $t+3$ incremented by $0.5 a_{t+1} +0.7  a_{t+2} + a_{t+3}$, and so on. 
+Delta adjustment allows imputations to differ sytematically from RBI methods and so provides a framework for performing sensitivity analysis around RBI assumptions.
+We follow the approach of the five macros. 
+Delta adjustment provides an increment which is added on to all values imputed after treatment discontinuation, but not to interim missing values. 
+It is specified by the options delta and dlag which we denote as $a_u$, a user-specified shift for time $u$, and $b_w$, a user-specified scaling multiplier that controls how the user-defined shift for time $u$ is applied to all times $\ge u$.
+Formally, any imputed value $Y^*_{z,t}$ is replaced by $Y^*_{z,t} + \sum_{u=t+1}^{u=\tmax} a_u b_{u-t}$.
+This means that values of delta are cumulated after treatment discontinuation.
+
+For example, consider an individual who discontinued treatment at the 2nd time point. 
+Under the setting \texttt{dlag=c(1,1,...,1)}, we take the vector of delta’s starting at the 3rd time point and add their cumulative sums to the imputed values. 
+Modifying dlag modifies this behaviour, so that the vector of delta’s starting at the 3rd time point is multiplied elementwise by the vector dlag. 
+A common increment of 3 at all time points after treatment discontinuation would be achieved by setting \texttt{delta=c(3,3,3,...)} and \texttt{dlag=c(1,0,0,...)}.
+An increment of 3 at the time point immediately after discontinuation that is halved at each subsequent time point is specified by  \texttt{delta=c(3,3,3,...)} and \texttt{dlag=c(1,-1/2,-1/4,...)}.
+
+%For example, if $b=(1,0.7, 0.5,...)$ then a participant who discontinues treatment just after visit $t$ has 
+%their imputed value at time $t+1$ incremented by $a_{t+1}$, 
+%their imputed value at time $t+2$ incremented by $0.7 a_{t+1} + a_{t+2}$, 
+%their imputed value at time $t+3$ incremented by $0.5 a_{t+1} +0.7  a_{t+2} + a_{t+3}$, and so on.
+
 Delta adjustment can apply either with RBI or the causal method and is always applied after imputation.
 
-\subsection{Interim missing values and covariates}
+\subsection{Interim missing values}
 
-[Kevin - to discuss]
+Interim missing values are values that are missing while the individual remains on treatment and are identified by being followed by later observed values.
+They are always imputed under MAR.
+Imputing interim missing values in a different way from post-discontinuation missing values is the procedure that is most likely to be appropriate in practice, since interim missing values that occur while on treatment are unlikely to follow the same pattern as post-discontinuation missing values that occur when off treatment.
+This differs from the behaviour of the Stata package \citep{Cro++16} which by default imputes interim missing values using the same procedure as post-discontinuation values, with MAR as an option.
+
+\subsection{Covariates}
+
+Covariates are handled ... [HOW? IAN THINK]
 
@@
 
-KEVIN TO DRAFT - INITIALLY TAKE FROM HELP.
+\subsection{Arguments} [KEVIN ARE WE ALLOWED TO CHANGE THE ORDER? BELOW MAKES MOST SENSE TO ME.]
 
-\subsection{Arguments}
+\begin{tabular}{llp{.8\linewidth}}\hline
+& Option & Description \\ \hline
+\multicolumn{3}{l}{\em Data options} \\
+& data 	&	 Dataset in long format, i.e.\ with one record per individual per time point. \\
+& covar 	&	 Covariates measured at or before baseline. Must be complete (no missing values).	\\
+& depvar 	&	 Dependent (outcome) variable.	\\
+& treatvar 	&	 Treatment group, coded 1,2,...	\\
+& idvar 	&	 Participant identifier.	\\
+& timevar 	&	 Time point for repeated measures. \\
+
+\multicolumn{3}{l}{\em Method options} \\
+& method 	&	 Reference-based imputation method: must be one of J2R, CIR, CR, LMCF, MAR, Causal.	\\
+& methodvar 	&	 Alternative to method option: variable in dataset specifying the imputation method  for each individual. \\
+& reference 	&	 Reference group for J2R, CIR, CR and Causal methods. \\
+& referencevar 	&	 Alternative to reference option: variable in dataset specifying the reference group for each individual. \\
+
+\multicolumn{3}{l}{\em Causal method options} \\
+& K0 	&	 Causal constant for use with Causal method.
+The treatment effect assumed at post-discontinuation times is K0 times the treatment effect at the time of discontinuation; if K1 $\ne$ 1, this is multiplied by a decaying term.
+\\
+& K1 	&	 Causal constant for use with Causal method. The treatment effect assumed at post-discontinuation times, implied by K0, decays exponentially by a factor K1 per time unit. \\
+
+\multicolumn{3}{l}{\em Delta method options}\\
+& delta 	&	 Optional vector of delta values to add onto imputed values (a’s in Rogers paper). Length must equal number of time points.\\
+& dlag 	&	 vector of delta values to add onto imputed values (non-mandatory) (b’s in Rogers paper) Length must equal number of time points. Default is c(1,1,1,...). \\
+
+\multicolumn{3}{l}{\em Computation options}\\
+& M 	&	 Number of imputations to be created. \\
+& seed 	&	 Seed value. Specify this so that a new run of the command will give the same imputed values.	\\
+
+\multicolumn{3}{l}{\em MCMC options} \\
+& prior 	&	 Prior for the variance-covariance matrix when fitting multivariate normal distributions: Jeffreys (default), uniform or ridge.	\\
+& burnin 	&	 Number of burn-in iterations when fitting multivariate normal distributions.	\\
+& bbetween 	&	 Number of iterations between imputed data sets when fitting multivariate normal distributions.	\\
+& mle 	&	 Logical option to use maximum likelihood parameter estimates instead of MCMC draw parameters. [NEED TO WRITE A COMMENT ABOUT THIS -- OR UNDOCUMENT IT?]	\\ \hline
+\end{tabular}
+
+\subsection{Algorithm}
+
+The program works through the following steps.
+\begin{enumerate}
+\item Set up a summary table based on treatment arm and missing data pattern (i.e.\ which timepoints are unobserved).
+\item Fit a multivariate normal distribution to each treatment arm using MCMC methods in package norm2 \citep{norm2}.
+\item Impute all interim missing values under a MAR assumption, looping over treatments and patterns.
+\item Impute post-discontinuation missing values under the user-specified assumption, looping over treatments and patterns.
+If methodvar and/or refvar is specified then the loop is also over teh values of these.
+\item Perform delta adjustment if specified.
+\item Repeat steps 2-5 M times and form into a single data frame.
+\end{enumerate}
+The baseline value of the outcome could be handled as an outcome, but this would allow a treatment effect at baseline [IAN CHECK]. We instead recommend handling it as a covariate.
+
+The program is based on Suzie Cro’s Stata program mimix.
+
 \subsection{Outputs}
+
+The M imputed data sets are output concatenated as one large dataset appended to
+the original unimputed dataset. [KEVIN IS IT A DATA SET, A DATA FRAME, OR OTHER?]
+
+The user can use the as.mids() function in the mice package to convert the output data  to mids data type and hence to perform analysis using Rubin's rules.
 
@@
 
-KEVIN TO DRAFT. Which data to use?
+KEVIN TO DRAFT PLEASE USING ASTHMA DATA.
 
@@
 \item doesn't cover other outcome types; e.g. methods have been proposed for recurrent events \citep{Keene++14}
+\item The suitability of the Rubin's rules standard errors has been debated \cite{ian:Carpenter_letter,Carpenter++14,Cro++19}...
 \item trials should be designed consistently with their estimand, so if the treatment-policy estimand is of interest then outcomes should be collected after treatment discontinuation.

### 2021-03-03 paper/paperRmimix.tex
Add text on covariates

@@
 
-\title{mimix: Reference-based imputation of missing data}
+\title{\texttt{mimix}: Reference-based imputation of missing data}
 \author{by Ian R White, Kevin McGrath, Matteo Quartagno, Suzie M Cro and James Carpenter}
@@
 Reference-based imputation is a multiple imputation technique which imputes quantitative outcome data that are missing after participant discontinuation of allocated treatment in a randomised trial.
-We present and describe an R package, mimix, for performing reference-based imputation, including a causal model variant, and we compare its implementation with that in SAS and Stata.
+We present and describe an R package, \texttt{mimix}, for performing reference-based imputation, including a causal model variant, and we compare its implementation with that in SAS and Stata.
 }
@@
 The causal model was previously specified for a two-arm trial and has been extended here for a multi-arm trial.
-The aim of this paper is to describe our implementation of mimix in R, to illustrate its use, and to describe its novel features by comparison with the implementations in SAS and Stata.
+The aim of this paper is to describe our implementation of \texttt{mimix} in R, to illustrate its use, and to describe its novel features by comparison with the implementations in SAS and Stata.
 
@@
 
+[ADD SOMETHING ON INDIVIDUALS WITH NO OBSERVED VALUES?]
+
 \subsection{Covariates}
 
-Covariates are handled ... [HOW? IAN THINK]
+Covariates are used to improve the plausibility of MAR assumptions and to improve the precision of imputed values \citep{ian:MItutorial}. 
+In \texttt{mimix}, they are included in the multivariate normal model fitting as part of the variable $Y_0$. [KEVIN IS THIS CORRECT? MIGHT THEY BE WRONGLY CARRIED FORWARD UNDER LMCF?]
+
+Missing values in covariates are not allowed in \texttt{mimix}.
+For these missing values in the randomised setting, single imputation methods are preferred, so mean imputation or the missing indicator method should be used before calling \texttt{mimix} \citep{ian:MIinRCTs}.
 
@@
 
-The program is based on Suzie Cro’s Stata program mimix.
+The program is based on Suzie Cro’s Stata program \texttt{mimix}.
 

### 2021-03-09 R/Runmimix.R
after matteos edits

@@ -1,47 +1,86 @@
 #' @title Main function for performing reference-based multiple imputation of longitudinal data  
 #' @description main wrapper for running mimix
-#' @details This is based on Suzie Cro's Stata program
-#' @details sets up a summary table based on missing data pattern- mg  mimix_group                                                                   
-#' @details reflects the pattern and treatment group configuration of the raw data
-#' @details then acts as a looping mechanism, norm2 is used as MCMC multivariate normal 
+#' @details The program works through the following steps
+#' @details 1. set up a summary table based on treatment arm and missing data pattern
+#' @details    (i.e. which timepoints are unobserved)
+#' @details 2. Fit a multivariate normal distribution to each treatment sarm using MCMC methods in package norm2
+#' @details 3. Impute all interim missing values  under a MAR assumption, looping ove treatments and patterns
+#' @details 4. Impute post-discontinuation missing values under the user-specified assumption,
+#' @details    looping over treatments and patterns (and over methodvar and referncevar if specified)
+#' @details 5. Perform delta-adjustment if specified
+#' @details 6. Repeat steps 2-5 M times and form into a single data frame 
+#' @details The baseline value of the outcome could be handed as an outcome, but this would allow a treatment effect at baseline
+#' @details We instead  recommend handling it as a covariate 
+#' @details The program is based on Suzie Cro's Stata program mimix
+#' @details The user can use the as.mids() function in the mice package to convert the output data to mids data type and hence 
+#' @details to perform analysis using Rubin's rules.    
 #' @export mimix
 #' @import mice
-#' @param data  Dataset in wide format
-#' @param covar Covariates - may include the baseline value of depvar. Must be complete (no missing values).
+#' @param data  Dataset in long format
+#' @param covar Covariates - baseline.  Must be complete (no missing values), enclose in quotes.
 #' @param depvar Dependent (outcome) variable
-#' @param treatvar Treatment group, coded 1,2,..
-#' @param idvar Participant id
[…]

### 2021-03-09 R/data.R

@@ -1,23 +1,30 @@
-#' asthma.
-#'
-#' A dataset containing  asthma trial data as used in the Stata mimix help file
-#'
-#'   @format A data frame containing 732 rows and 5 columns
+#' Sample data: asthma trial
+#' A data set containing  asthma trial data as used in the Stata mimix help file
+#' The primary outcome variable is fev,  measured at 2,4,8,12 weeks  
+#' @format A data frame containing 732 rows and 5 columns
 #'   \describe{
-#'   \item{id}{}
+#'   \item{id} {patient identifier}{}
 #'   \item{time}{}
 #'   \item{treat}{}
-#'   \item{base}{}
-#'   \item{fev}{}
+#'   \item{base}{covariate}{}
+#'   \item{fev}{dependent variable}{}
 #'   }
+#' @examples
+#' \dontrun{
+#' impJ2Rridge<-(mimix(data=asthma,covar=c("base"),depvar=fev,treatvar=treat,idvar=id,timevar=time,
+#'     method="J2R",reference=1,delta=c(0.5,0.5,1,1 ),M=5,seed=101,prior=ridge,burnin=1000)
+#' library(mice)     
+#' fit<-with(data= as.mids(impJ2Rridge),lm(fev.12~treat))
+#' summary(pool(fit))    
+#' }  
 "asthma"
 
 
 
-#' antidepressant.
-#'  A data set containing  antidepressant trial data as described in paper by White,Royes,Best
-#'
-#' @format  dataframe containing 688 rows and 14 columns
+#' Sample data: antidepressant trial
+#' A data set containing antidepressant trial data as described in paper by White, Royes, Best (2019)
+#' The primary outcome is HAMD17.TOTAL measured at visit number 4,5,6,7. 
+#' @format  A data frame containing 688 rows and 14 columns
 #'  \describe{
 #'  \item{PATIENT.NUMBER}{}
 #'  \item{HAMA.TOTAL}{}
@@ -27,23 +34,30 @@
 #'  \item{TREATMENT.NAME}{}
 #'  \item{PATIENT.SEX}{}
 #'  \item{POOLED.INVESTIGATOR}{}
-#'  \item{basval}{}
-#'  \item{HAMD17.TOTAL}{}
+#'  \item{basval}{covariate}{}
+#'  \item{HAMD17.TOTAL}{dependent variable}{}
 #'  \item{change}{}
 #'  \item{miss_flag}{}
[…]

### 2021-03-09 R/mimi_package.R

@@ -1,18 +1,51 @@
-#' mimix: A package porting the Stata mimix command  
+#' @details mimix: Comparisons with Stata and SAS   
+#' 
+#' @details # Comparison with Stata:
+#' @details This mimix is based on the Stata version and has similar functionality while adding
+#' @details the causal method and delta adjustment. 
+#' @details As with the Stata version the input data requires the longitudinal input data in
+#' @details long format with one record per individual at each timepoint.
+#' @details The program differs in how interim missing cases - those cases which 
+#' @details have a missing measurement at a timepoint previous to a later observed measurement - are treated.        
+#' @details Under Stata by default, the interim missing are treated the same as for the post-discontinuation
+#' @details missing unless the interim option is explicitly used. 
+#' @details Here the interims are treated as under MAR, the post-discontinuations then imputed under
+#' @details the specified method. There is no interim option as there is in Stata.
+#' @details Unlike Stata an option is supplied whereby the prior used in the MCMC draws can be changed from the 
+#' @details default jeffreys (as in Stata) to either the ridge or uniform  
+#' 
+#' @details # Comparison with SAS:
+#' 
+#' @details Whilst this program is based on the Stata program, the latter is an adaptation of the SAS macro miwithd,
+#' @details written by james Roger, subsequently updated to the Five_Macros suite of macros
+#' @details This program uses the same approach for the delta adjustment as described in the Five_macros, 
+#' @details in comparing outputs from our program with the Five_macros it is to be noted that interaction between treatment and covariates
+#' @details is not allowed in the SAS macros, and comparisons are only valid for example in testing the Causal model by specifically not
+#' @details not using the covbytime and catcovbytime options in the Five_macros 
+#' @details Not using these options also means that the LMCF method can be compared with either ALMCF or OLMCF in the Five_macros.
+#' @details When there is no observed data (common in the acupuncture data) the first mean is used in Stata,
+#' @details a warning is given in the Five_macros     
 #'
-#' The mimix package provides the functionality of the Stata package plus
-#' delta and causal methods
-#' 
-#' @section Comparison with Stata:
-#' The mimix functions add the causal method and delta extra to the Stata version  ...
-#' 
-#' @section Comparison with SAS:
[…]

### 2021-03-09 R/proprocess.R

@@ -16,6 +16,8 @@
 
 
 #preprocess data for group method
+#edits to get("data" 05/01/21)
+
 preprodata<- function(data,covar,depvar,treatvar,idvar,timevar,M,reference,method=NULL)  {
   #extract relevant vars
    
@@ -27,11 +29,11 @@ preprodata<- function(data,covar,depvar,treatvar,idvar,timevar,M,reference,metho
     # fevdata<-dplyr::select(get(data),idvar,depvar,timevar)
   ##2311 make sure .id is id ! 
   #browser()
-    fevdata<- get(data)[c(idvar,covar,depvar,timevar,treatvar)]
+    fevdata<- get("data")[c(idvar,covar,depvar,timevar,treatvar)]
   # extract covar cols 1 row per id to merge onto the wide data
-    uniqdat<-unique(get(data)[c(idvar,covar,treatvar)])
-    ntreatcol<- get(data)[c(treatvar)]
-    ntimecol<- get(data)[c(timevar)]
+    uniqdat<-unique(get("data")[c(idvar,covar,treatvar)])
+    ntreatcol<- get("data")[c(treatvar)]
+    ntimecol<- get("data")[c(timevar)]
 
 
 
@@ -79,7 +81,7 @@ preprodata<- function(data,covar,depvar,treatvar,idvar,timevar,M,reference,metho
 
   # process no covars
 
-  testfevdata<- get(data)[c(idvar,depvar,timevar)]
+  testfevdata<- get("data")[c(idvar,depvar,timevar)]
   sts4dummy<-stats::reshape(testfevdata,v.names = depvar,timevar = timevar,idvar=idvar,direction="wide")
   STSdummy<- apply(sts4dummy[,grepl(depvar,names(sts4dummy))],MARGIN=2,function(x) ifelse(!is.na(x),0,1))
 
@@ -224,11 +226,11 @@ preprodata<- function(data,covar,depvar,treatvar,idvar,timevar,M,reference,metho
 preproIndivdata<- function(data,covar,depvar,treatvar,idvar,timevar,M,reference=NULL,method=NULL,methodvar,referencevar)  {
  #browser()
   #check covars complete
-  stopifnot(sum(is.na(get(data)[,covar]))==0)
+  stopifnot(sum(is.na(get("data")[,covar]))==0)
   #tryCatch(stopifnot(sum(is.na(mxdata[,covar]))!=0,error=stop("Error: not all covariates are complete !!")))
[…]

### 2021-03-10 paper/paperRmimix.tex
Add comparison and example to paper

@@
 \abstract{
-Reference-based imputation is a multiple imputation technique which imputes quantitative outcome data that are missing after participant discontinuation of allocated treatment in a randomised trial.
-We present and describe an R package, \texttt{mimix}, for performing reference-based imputation, including a causal model variant, and we compare its implementation with that in SAS and Stata.
+In some randomised trials, outcome data are missing after participant discontinuation of allocated treatment.
+Reference-based imputation is a multiple imputation technique for quantitative outcome data in this setting: it constructs the imputation model using parameters from a reference group.
+We present and describe an R package, \texttt{mimix}, for performing reference-based imputation.
+The package also includes a causal model variant and imputation of data missing before discontinuation of allocated treatment.
+We compare its implementation with that in SAS and Stata.
 }
@@
 
-The implicit assumptions behind this approach were explored by \citet{ian:RBIcausal}, who proposed a causal model in which the treatment effect after discontinuation is a specified multiple of the treatment effect at the point of discontinuation.
+The implicit assumptions behind this approach were explored by \citet{ian:RBIcausal}, who proposed a causal model in which the treatment effect after discontinuation is a specified fraction of the treatment effect at the point of discontinuation.
 They showed that J2R, CR and CIR are the special cases of this causal model in which the treatment effect disappears, decays or is maintained after treatment discontinuation, while LMCF and MAR are not special cases of the causal model.
@@
 
-\subsection{Setting} [nicked from \citet{ian:RBIcausal}]
-We assume that quantitative outcome measurements are scheduled at baseline and at $\tmax$ occasions after randomisation. 
-Let $Z$ be the random variable for the participant's randomised treatment arm, let $Z=z$ denote an arm in which we want to impute missing values, and $Z=r$ denote the reference arm.
+\subsection{Setting} 
+Following \citet{ian:RBIcausal},
+we assume that quantitative outcome measurements are scheduled at baseline and at $\tmax$ occasions after randomisation. 
+Let $Z$ be the random variable for the participant's randomised treatment arm, where $Z=z$ denotes an arm in which we want to impute missing values, and $Z=r$ denotes the reference arm.
 Let $Y_t$ be the random variable for the participant's outcome at visit $t=0,...,\tmax$. 
-It is convenient to define 
-$\boldsymbol{Y}_{\pret}=(Y_0,\ldots,Y_t)$, the vector of all outcomes up to  and including visit $t$; 
-$\boldsymbol{Y}_{\postt}=(Y_{t+1},\ldots,Y_\tmax)$, the vector of all outcomes after visit $t$; 
+It is convenient to define row vectors
+$\boldsymbol{Y}_{\pret}=(Y_0,\ldots,Y_t)$ containing all outcomes up to  and including visit $t$; 
+$\boldsymbol{Y}_{\postt}=(Y_{t+1},\ldots,Y_\tmax)$ containing all outcomes after visit $t$; 
 and
-$\boldsymbol{Y}=(Y_0,\ldots,Y_\tmax)$, the vector of all outcomes.
+$\boldsymbol{Y}=(Y_0,\ldots,Y_\tmax)$, containing all outcomes.
 Let $D$ be the random variable for the participant's last visit prior to discontinuing treatment, so $D=0,...,\tmax$.
@@
 
-\item For each treatment arm $z$, draw a mean vector $\boldsymbol{\mu}_z^*$ and variance-covariance matrix from the
-    posterior distribution $\boldsymbol{\Sigma}_z^*$.
+\item For each treatment arm $z$, draw a mean vector $\boldsymbol{\mu}_z^*$ and variance-covariance matrix $\boldsymbol{\Sigma}_z^*$ from the posterior distribution.
 
-\item For each treatment arm  $z$ and each possible treatment discontinuation visit $t$, use the drawn values to build a hypothetical  joint  distribution of 
-the outcomes $\boldsymbol{Y}_{\pret}$ up to time $t$ 
+\item For each treatment arm  $z$ and each observed treatment discontinuation visit $t$, use the drawn values to build a hypothetical  joint  distribution of 
+the outcomes $\boldsymbol{Y}_{\pret}$ up to visit $t$ 
 and 
-the outcomes $\boldsymbol{Y}_{\postt}$ after time  $t$, 
+the outcomes $\boldsymbol{Y}_{\postt}$ after visit  $t$, 
 using one of the methods described below.
 Thus a MVN distribution is built for $\boldsymbol{Y}|Z=z,D=t$.
+We define the means
+$\boldsymbol{\mu}_{z,\pret}$,
+$\boldsymbol{\mu}_{z,\postt}$
+to be the parts of $\boldsymbol{\mu}_{z}$ up to and after visit $t$ respectively.
+
 Five methods are mainly distinguished by their choice of mean: 
@@
 
-The variance matrices are constructed so that the regression coefficient matrix and conditional variance matrix of the potential outcomes after visit  $t$ given those before visit $t$ are taken from arm $z$ for MAR and LMCF, and from arm $r$ for J2R, CIR and CR
+The variance matrices are constructed so that the regression coefficient matrix and conditional variance matrix of the potential outcomes after visit  $t$, given those before visit $t$, are taken from arm $z$ for MAR and LMCF, and from arm $r$ for J2R, CIR and CR
 \citet{Carpenter++13,ian:RBIcausal}.
-[NOTE we didn't code the \emph{RBI alternative} approach of \citet{ian:RBIcausal} that instead uses $\boldsymbol{\beta}_t(\tmax)$ and $\boldsymbol{\Omega}_t(\tmax)$ for all RBI methods.]
 
@@
 In particular, $Y_t(z,0)$ is the potential outcome if never treated: it is the same for all $z$ and is written $Y_t(0)$. Similarly $Y_t(z,\tmax)$ is the potential outcome if always treated with $z$.
-We define 
-$\boldsymbol{Y}_{\pret}(z,s)$, $\boldsymbol{Y}_{\postt}(z,s)$ and $\boldsymbol{Y}(z,s)$ as before.
 We let 
-$\mu_t(z,s) = \E[Y_t(z,s)]$, the mean of the potential outcome at visit $t$ if active treatment $z$ were received for $s$ periods only. Similarly we define 
-$\boldsymbol{\mu}_\pret(z,s)$, $\boldsymbol{\mu}_\postt(z,s)$ and $\boldsymbol{\mu}(z,s)$.
+$\mu_t(z,s) = \E[Y_t(z,s)]$, the mean of the potential outcome at visit $t$ if active treatment $z$ were received for $s$ periods only. 
+We define row vectors
+$\boldsymbol{Y}_{\pret}(z,s)$, $\boldsymbol{Y}_{\postt}(z,s)$, $\boldsymbol{Y}(z,s)$,
+$\boldsymbol{\mu}_\pret(z,s)$, $\boldsymbol{\mu}_\postt(z,s)$ and $\boldsymbol{\mu}(z,s)$
+ as before.
 
@@
 
-[AT PRESENT $(k_0, k_1)$ ARE FIXED BUT IT MAY BE POSSIBLE TO LET THEM DEPEND ON ARM.
-THIS MAKES SENSE FOR A TRIAL WITH MORE THAN 2 ARMS.
-WOULD PROBABLY CODE THEM AS A VARIABLE IN DATA SET SO COULD LET THEM VARY BY OTHER FACTORS E.G.
-REASON?]
+[At present $(k_0, k_1)$ are fixed but it may be possible to let them depend on arm. 
+This makes sense for a trial with more than 2 arms. 
+Would probably code them as a variable in data set so could let them vary by any individual factors e.g. reason for missing data]
 
@@
 
+\subsection{Interim missing values}
+
+Interim missing values are values that are missing while the individual remains on treatment and are identified by being followed by later observed values. 
+In the RBI algorithm and causal model, applying the constructed imputation distribution to missing values in $\boldsymbol{Y}_{\pret}$ implies imputation under MAR.
+We consider MAR imputation of interim missing values to be the procedure that is most likely to be appropriate in practice, since interim missing values that occur while on treatment are unlikely to follow the same pattern as post-discontinuation missing values that occur when off treatment.
+This differs from the behaviour of the Stata package \citep{Cro++16} which by default imputes interim missing values using the same procedure as post-discontinuation values, with MAR as an option.
+
+[ADD SOMETHING ON INDIVIDUALS WITH NO OBSERVED VALUES]
+
 \subsection{Delta adjustment}
@@
 Delta adjustment provides an increment which is added on to all values imputed after treatment discontinuation, but not to interim missing values. 
-It is specified by the options delta and dlag which we denote as $a_u$, a user-specified shift for time $u$, and $b_w$, a user-specified scaling multiplier that controls how the user-defined shift for time $u$ is applied to all times $\ge u$.
+It is specified by two user-specified $T$-dimensional row vectors $\boldsymbol{a}$ and $\boldsymbol{b}$,
+where $a_u$ is a shift for time $u$, 
+and $b_w$ is a scaling multiplier that controls how the user-defined shift for time $u$ is applied at time $u+w$.
 Formally, any imputed value $Y^*_{z,t}$ is replaced by $Y^*_{z,t} + \sum_{u=t+1}^{u=\tmax} a_u b_{u-t}$.
-This means that values of delta are cumulated after treatment discontinuation.
+This means that elements of $\boldsymbol{a}$  are cumulated after treatment discontinuation.
 
 For example, consider an individual who discontinued treatment at the 2nd time point. 
-Under the setting \texttt{dlag=c(1,1,...,1)}, we take the vector of delta’s starting at the 3rd time point and add their cumulative sums to the imputed values. 
-Modifying dlag modifies this behaviour, so that the vector of delta’s starting at the 3rd time point is multiplied elementwise by the vector dlag. 
-A common increment of 3 at all time points after treatment discontinuation would be achieved by setting \texttt{delta=c(3,3,3,...)} and \texttt{dlag=c(1,0,0,...)}.
-An increment of 3 at the time point immediately after discontinuation that is halved at each subsequent time point is specified by  \texttt{delta=c(3,3,3,...)} and \texttt{dlag=c(1,-1/2,-1/4,...)}.
+Under the setting $b_w=1$ for all $w$, we take the vector $\boldsymbol{a}$ starting at the 3rd element and add the cumulative sums to the imputed values. 
+Modifying $\boldsymbol{b}$ modifies this behaviour, so that the vector of $\boldsymbol{a}$ starting at the 3rd time point is multiplied elementwise by the vector $\boldsymbol{b}$. 
+A common increment of 3 at all time points after treatment discontinuation would be achieved by setting $\boldsymbol{a}=(3,3,3,...)$ and $\boldsymbol{b}=(1,0,0,...)$.
+An increment of 3 at the time point immediately after discontinuation that is halved at each subsequent time point is specified by  
+$\boldsymbol{a}=(3,3,3,...)$ and $\boldsymbol{b}=(1,-1/2,-1/4,...)$.
 
@@
 
-\subsection{Interim missing values}
-
-Interim missing values are values that are missing while the individual remains on treatment and are identified by being followed by later observed values.
-They are always imputed under MAR.
-Imputing interim missing values in a different way from post-discontinuation missing values is the procedure that is most likely to be appropriate in practice, since interim missing values that occur while on treatment are unlikely to follow the same pattern as post-discontinuation missing values that occur when off treatment.
-This differs from the behaviour of the Stata package \citep{Cro++16} which by default imputes interim missing values using the same procedure as post-discontinuation values, with MAR as an option.
-
-[ADD SOMETHING ON INDIVIDUALS WITH NO OBSERVED VALUES?]
-
 \subsection{Covariates}
@@
 Covariates are used to improve the plausibility of MAR assumptions and to improve the precision of imputed values \citep{ian:MItutorial}. 
-In \texttt{mimix}, they are included in the multivariate normal model fitting as part of the variable $Y_0$. [KEVIN IS THIS CORRECT? MIGHT THEY BE WRONGLY CARRIED FORWARD UNDER LMCF?]
+In \texttt{mimix}, they are included in the multivariate normal model fitting as part of the variable $Y_0$. 
+[KEVIN IS THIS CORRECT? MIGHT THEY BE WRONGLY CARRIED FORWARD UNDER LMCF?]
 
@@
 
-\subsection{Arguments} [KEVIN ARE WE ALLOWED TO CHANGE THE ORDER? BELOW MAKES MOST SENSE TO ME.]
+\subsection{Arguments}
 
@@
 \multicolumn{3}{l}{\em Delta method options}\\
-& delta 	&	 Optional vector of delta values to add onto imputed values (a’s in Rogers paper). Length must equal number of time points.\\
-& dlag 	&	 vector of delta values to add onto imputed values (non-mandatory) (b’s in Rogers paper) Length must equal number of time points. Default is c(1,1,1,...). \\
+& delta 	&	 Optional vector $\boldsymbol{a}$ of values to add onto imputed values as described above under ``Delta adjustment''. Length must equal number of time points.\\
+& dlag 	&	 Optional vector $\boldsymbol{b}$ of values to modify the values in delta. Length must equal number of time points. \\
 
@@
 & bbetween 	&	 Number of iterations between imputed data sets when fitting multivariate normal distributions.	\\
-& mle 	&	 Logical option to use maximum likelihood parameter estimates instead of MCMC draw parameters. [NEED TO WRITE A COMMENT ABOUT THIS -- OR UNDOCUMENT IT?]	\\ \hline
+& mle 	&	 Logical option to use maximum likelihood parameter estimates instead of MCMC draw parameters. 
+[NEED TO WRITE A COMMENT ABOUT THIS -- KEVIN, OR CAN WE UNDOCUMENT IT?]	\\ \hline
 \end{tabular}
@@
 \end{enumerate}
-The baseline value of the outcome could be handled as an outcome, but this would allow a treatment effect at baseline [IAN CHECK]. We instead recommend handling it as a covariate.
+The baseline value of the outcome could be handled as an outcome or as a covariate. 
+[Which do we recommend?]
 
@@
 The M imputed data sets are output concatenated as one large dataset appended to
-the original unimputed dataset. [KEVIN IS IT A DATA SET, A DATA FRAME, OR OTHER?]
+the original unimputed dataset.
+[KEVIN IS IT A DATA SET, A DATA FRAME, OR OTHER?]
 
-The user can use the as.mids() function in the mice package to convert the output data  to mids data type and hence to perform analysis using Rubin's rules.
+The user can use the \texttt{as.mids()} function in the \texttt{mice} package to convert the output data  to \texttt{mids} data type and hence to perform analysis using Rubin's rules.
 
@@
 
-\section{Examples}
+\section{Example: asthma trial data}
 
-KEVIN TO DRAFT PLEASE USING ASTHMA DATA.
+These data are taken from \citet{Cro++16} and represent 2 arms of a 5-arm trial. 
+They contain 183 patients with asthma who were randomised to active treatment or placebo. 
+The primary outcome is FEV at 12 weeks, and FEV is also measured at baseline and at 2, 4, and 8 weeks of follow-up.
+The data contain 732 records (one per patient per follow-up visit) and 5 variables: 
+id, a patient identifier; 
+time, the visit number; 
+treat, the randomised group (1=placebo, 2=active); 
+base, the baseline FEV, to be regarded as a covariate;
+and fev, the outcome variable.
+
+The data are missing after patients discontinue their randomised treatment.
+A treatment policy estimand is therefore estimated using reference-based imputation.
+For example, J2R is implemented as follows
+\begin{verbatim}
+asthmaJ2R <- mimix(data='asthma', covar=base, depvar=fev, treatvar=treat, idvar=id, timevar=time,
+    method='J2R', reference=1, M=5, seed=101, prior=ridge, burnin=1000)
+\end{verbatim}
+
+[KEVIN, IT SEEMS COVAR MUST BE STRING BUT DEPVAR A VARNAME, IS THIS CORRECT? NEEDS A COMMENT?]
+
+Note that the imputed data are in wide format. 
+[KEVIN / ALL - I THINK THIS IS A PROBLEM!?]
+The primary outcome can be analysed using an analysis of covariance by
+\begin{verbatim}
+library(mice)     
+fit <- with(data= as.mids(asthmaJ2R), lm(fev.12 ~ treat + base))
+summary(pool(fit))    
+\end{verbatim}
+
+[GIVE RESULTS]
 
@@
 
-KEVIN TO DRAFT -- THESE ARE JUST MY NOTES. Would a table be suitable?
+[TO DISCUSS -- TEXT OR TABLE? BOTH ARE INCLUDED BELOW.]
 
-\begin{enumerate}
-\item flavours of LMCF in 5 macros
-\item handling participants with no observed outcomes
-\item how baseline covariates are modelled: SAS, can interact with time; Stata, R, can't
-\item handling of interim missing values
-\item anything to say about algorithm differences? prior choice?
-\end{enumerate}
+\subsection{Comparison with Stata}
+This mimix is based on the Stata version and has similar functionality while adding
+the causal method and delta adjustment. 
+As with the Stata version the input data requires the longitudinal input data in
+long format with one record per individual at each timepoint.
+
+The program differs in how interim missing cases -- those cases which 
+have a missing measurement at a timepoint previous to a later observed measurement -- are treated.        
+Under Stata by default, the interim missing are treated the same as for the post-discontinuation
+missing unless the interim option is explicitly used. 
+Here the interims are treated as under MAR, the post-discontinuations then imputed under
+the specified method. There is no interim option as there is in Stata.
+
+Unlike Stata an option is supplied whereby the prior used in the MCMC draws can be changed from the 
+default jeffreys (as in Stata) to either the ridge or uniform.
+
+\subsection{Comparison with SAS}
+
+Whilst this program is based on the Stata program, the latter is an adaptation of the SAS macro miwithd,
+written by James Roger, subsequently updated to the Five\_Macros suite of macros.
+This program uses the same approach for the delta adjustment as described in the Five\_Macros.
+
+In comparing outputs from our program with the Five\_Macros it is to be noted that interaction between treatment and covariates
+is not allowed in the SAS macros, and comparisons are only valid for example in testing the Causal model by specifically not
+not using the covbytime and catcovbytime options in the Five\_Macros.
+
+Not using these options also means that the LMCF method can be compared with either ALMCF or OLMCF in the Five\_Macros: these are the same in the mimix setting with no covariate interacting with time.
+
+When there is no observed data (common in the acupuncture data) the first mean is used in Stata,
+a warning is given in the Five\_Macros.
+
+\subsection{summary table}
+
+\begin{tabular}{p{.2\textwidth}p{.2\textwidth}p{.2\textwidth}p{.2\textwidth}}\hline
+	&	Five macros in SAS	&	mimix in Stata	&	mimix in R	\\\hline
+{\em Functionality}	&		&		&		\\
+RBI	&	yes	&	yes	&	yes	\\
+Causal model	&	yes	&	no	&	yes	\\
+Delta method	&	yes	&	no	&	yes	\\\hline
+{\em Options}	&		&		&		\\
+Interim missing	&	MAR	&	RBI or MAR	&	MAR	\\
+MCMC options	&	seed and thinning (=burnin)	&	seed, burnin and thinning	&	seed, burnin, thinning and prior	\\
+Covariates interacting with treatment	&	never	&	always	&	always	\\
+Covariates interacting with time	&	optional	&	never	&	never	\\
+LMCF	&	several flavours: ALMCF, OLMCF, AFCMCF, OFCMCF	&	ALMCF only?	&	ALMCF only?	\\
+No observed data	&	Warning	&	First mean is used	&	?	\\\hline
+\end{tabular}
 
@@
 
-\section{Limitations and discussion}
+\section{Limitations and discussion: sketch only so far}
 
 \begin{itemize}
-\item flexible package
+\item Flexible package
 \item doesn't cover other outcome types; e.g. methods have been proposed for recurrent events \citep{Keene++14}
-\item The suitability of the Rubin's rules standard errors has been debated \cite{ian:Carpenter_letter,Carpenter++14,Cro++19}...
-\item trials should be designed consistently with their estimand, so if the treatment-policy estimand is of interest then outcomes should be collected after treatment discontinuation.
+\item The suitability of the Rubin's rules standard errors has been debated \citep{ian:Carpenter_letter,Carpenter++14,Cro++19}
+\item Trials should be designed consistently with their estimand, so if the treatment-policy estimand is of interest then outcomes should be collected after treatment discontinuation.
 Analysis options for this setting are still in development: options include including treatment discontinuation time in the model and imputing under MAR [ref Tom at GSK]; using RBI or causal model and reserving the post-discontinuation data for model checking; or using RBI or causal model and using the post-discontinuation data to estimate model parameters such as $K_{z,t}$.
+\item We didn't code the \emph{RBI alternative} approach of \citet{ian:RBIcausal} that instead uses $\boldsymbol{\beta}_t(\tmax)$ and $\boldsymbol{\Omega}_t(\tmax)$ for all RBI methods.
+
 \end{itemize}

### 2021-03-12 R/Runmimix.R
V0.0.12

@@ -22,7 +22,7 @@
 #' @param treatvar Treatment group, can be character
 #' @param idvar Participant identifier.
 #' @param timevar Time point for repeated measure
-#' @param method Reference-based imputation method: must be J2R, CR, CIR, MAR, Causal or LMCF,enclose in quotes. 
+#' @param method Reference-based imputation method: must be J2R, CR, CIR, MAR, Causal or LMCF, enclose in quotes. 
 #' @param reference  Reference group for J2R, CIR, CR methods
 #' @param methodvar column in data-set specifying individual method, enclose in quotes.
 #' @param referencevar column in data-set specifying reference group as forindividual method, enclose in quotes.
@@ -41,18 +41,22 @@
 #' @examples
 #' \dontrun{
 #' "performing jump to reference with treatment reference arm 1 on asthma trial data"  
-#' mimixout<-mimix(data=asthma,covar=c(base),depvar=fev,treatvar=treat,idvar=id,timevar=time,method="J2R",reference=1,M=5,seed=54321)
+#' mimixout<-mimix(data=asthma,covar=c(base),depvar=fev,treatvar=treat,idvar=id,timevar=time,
+#'  method="J2R", reference=1,M=5,seed=54321)
 #' library(mice)
-#' "Fitting regression model to find treatment effects using Rubin's rules by treating output data frame as.mids() object "
+#' "Fitting regression model to find treatment effects using Rubin's rules by 
+#'      treating output data frame as.mids() object "
 #' fit<-with(data= as.mids(mimixout),expr = lm(fev.12~treat+base))
 #' summary(pool(fit))
-#' mimix(data=acupuncture,covar= c(head_base),depvar=head,treatvar=treat,idvar=id,timevar=time,method="CIR",reference=1,M=5,seed=54321,
+#' mimix(data=acupuncture,covar= c(head_base),depvar=head,treatvar=treat,idvar=id,
+#'  timevar=time,method="CIR",reference=1,M=5,seed=54321,
 #'   prior=jeffreys,burnin=1000)
 #' }
 
 # v0.0.12
[…]

### 2021-03-12 R/data.R

@@ -1,30 +1,36 @@
 #' Sample data: asthma trial
+#' 
 #' A data set containing  asthma trial data as used in the Stata mimix help file
-#' The primary outcome variable is fev,  measured at 2,4,8,12 weeks  
+#' The primary outcome variable is fev, measured at 2,4,8,12 weeks
+#'
+#'
 #' @format A data frame containing 732 rows and 5 columns
 #'   \describe{
-#'   \item{id} {patient identifier}{}
+#'   \item{id}{patient identifier}
 #'   \item{time}{}
 #'   \item{treat}{}
-#'   \item{base}{covariate}{}
-#'   \item{fev}{dependent variable}{}
+#'   \item{base}{covariate}
+#'   \item{fev}{dependent variable}
 #'   }
 #' @examples
 #' \dontrun{
-#' impJ2Rridge<-(mimix(data=asthma,covar=c("base"),depvar=fev,treatvar=treat,idvar=id,timevar=time,
+#'  impJ2Rridge<-(mimix(data=asthma,covar=c("base"),depvar=fev,treatvar=treat,idvar=id,timevar=time,
 #'     method="J2R",reference=1,delta=c(0.5,0.5,1,1 ),M=5,seed=101,prior=ridge,burnin=1000)
-#' library(mice)     
-#' fit<-with(data= as.mids(impJ2Rridge),lm(fev.12~treat))
-#' summary(pool(fit))    
-#' }  
+#'  library(mice)     
+#'  fit<-with(data= as.mids(impJ2Rridge),lm(fev.12~treat))
+#'  summary(pool(fit))    
+#' }     
 "asthma"
 
 
 
 #' Sample data: antidepressant trial
-#' A data set containing antidepressant trial data as described in paper by White, Royes, Best (2019)
-#' The primary outcome is HAMD17.TOTAL measured at visit number 4,5,6,7. 
-#' @format  A data frame containing 688 rows and 14 columns
+#' 
+#' A data set containing antidepressant trial data as described in paper 
+#'   by White,Royes,Best (2019)
+#' The primary outcome is HAMD17.TOTAL measured at visit number 4,5,6,7.  
+#'
+#' @format  dataframe containing 688 rows and 14 columns
 #'  \describe{
 #'  \item{PATIENT.NUMBER}{}
 #'  \item{HAMA.TOTAL}{}
[…]

### 2021-03-12 R/mimix_comparison.R

@@ -1,0 +1,56 @@
+#' mimix: Comparisons with Stata and SAS   
+#' 
+#' @description 
+#' mimix is available in Stata whils th Five_macros suite does similar in SAS  
+#' 
+#' @section  Comparison with Stata:
+#' This mimix is based on the Stata version and has similar functionality while adding
+#' the causal method and delta adjustment. 
+#' As with the Stata version the input data requires the longitudinal input data in
+#' long format with one record per individual at each timepoint.
+#' The program differs in how interim missing cases - those cases which 
+#' have a missing measurement at a timepoint previous to a later observed measurement - are treated.        
+#' Under Stata by default, the interim missing are treated the same as for the post-discontinuation
+#' missing unless the interim option is explicitly used. 
+#' Here the interims are treated as under MAR, the post-discontinuations then imputed under
+#' the specified method. There is no interim option as there is in Stata.
+#' Unlike Stata an option is supplied whereby the prior used in the MCMC draws can be changed from the 
+#' default jeffreys (as in Stata) to either the ridge or uniform  
+#' 
+#' 
+#' @section Comparison with SAS:
+#' 
+#' Whilst this program is based on the Stata program, the latter is an adaptation of the SAS macro miwithd,
+#' written by james Roger, subsequently updated to the Five_Macros suite of macros
+#' This program uses the same approach for the delta adjustment as described in the Five_macros, 
+#' in comparing outputs from our program with the Five_macros it is to be noted that interaction between treatment and covariates
+#' is not allowed in the SAS macros, and comparisons are only valid for example in testing the Causal model by specifically not
+#' not using the covbytime and catcovbytime options in the Five_macros 
+#' Not using these options also means that the LMCF method can be compared with either ALMCF or OLMCF in the Five_macros.
+#' When there is no observed data (common in the acupuncture data) the first mean is used in Stata,
+#' a warning is given in the Five_macros     
+#'
+#'  
+#'  
+#' @references 
+#'     Cro s, Morris T, Kenward G,Carpenter Joshttps://www.ncbi.nlm.nih.gov/pmc/articles/PMC5796638/
+#'     White I, Royes J, Best N, https://arxiv.org/abs/1705.04506
+#'     Roger J,
+#'     URL: https://www.lshtm.ac.uk/research/centres-projects-groups/missing-data#sensitivity-analysis, 
+#' @docType package
+#' @name mimix_Comparison 
[…]

### 2021-03-12 R/Runmimix.R
edit example quotes

@@ -41,14 +41,14 @@
 #' @examples
 #' \dontrun{
 #' "performing jump to reference with treatment reference arm 1 on asthma trial data"  
-#' mimixout<-mimix(data=asthma,covar=c(base),depvar=fev,treatvar=treat,idvar=id,timevar=time,
+#' mimixout<-mimix(data=asthma,covar=c("base"),depvar=fev,treatvar=treat,idvar=id,timevar=time,
 #'  method="J2R", reference=1,M=5,seed=54321)
 #' library(mice)
 #' "Fitting regression model to find treatment effects using Rubin's rules by 
 #'      treating output data frame as.mids() object "
 #' fit<-with(data= as.mids(mimixout),expr = lm(fev.12~treat+base))
 #' summary(pool(fit))
-#' mimix(data=acupuncture,covar= c(head_base),depvar=head,treatvar=treat,idvar=id,
+#' mimix(data=acupuncture,covar= c("head_base"),depvar=head,treatvar=treat,idvar=id,
 #'  timevar=time,method="CIR",reference=1,M=5,seed=54321,
 #'   prior=jeffreys,burnin=1000)
 #' }

### 2021-03-12 R/data.R

@@ -84,7 +84,7 @@
 #' \dontrun{
 #'  impCausalref1 <- mimix(data=acupuncture,covar=c("head_base","sex"),depvar=head,treatvar=treat,
 #'        idvar=id,timevar=time,
-#'        method=Causal,reference=1,K0=1,K1=0.5,M=5,seed=54321)
+#'        method="Causal",reference=1,K0=1,K1=0.5,M=5,seed=54321)
 #'  library(mice)         
 #'  fit<-with(as.mids(impCausalref1), lm(head.12~treat+head_base+sex))  
 #'  summary(pool(fit))    

### 2021-03-13 paper/paperRmimix.tex
Update paper after discussion with Kevin. Explore how wholly-missing observatins are handled in Stata and R

@@
 
-[ADD SOMETHING ON INDIVIDUALS WITH NO OBSERVED VALUES]
+\subsection{Individuals with no observed values}
+
+[TEXT FOR DISCUSSION -- NEED TO WORK OUT HOW TO HANDLE THESE DISCREPANCIES]
+
+Individuals with no observed values are easily imputed under MAR, but for the other methods they are currently a problem that we need to resolve.
+
+We first consider J2R and CR. 
+These methods should construct the imputation distribution wholly from the reference arm. 
+They therefore correspond to MAR imputation if the wholly unobserved individual was in the reference arm. 
+Both the Stata and R programs do this correctly for an individual in the reference arm, but for an individual in a non-reference arms, J2R and CR do not agree with each other.
+
+We next consider CIR and LMCF. These are defined in terms of the last observed value, so they are formally undefined. 
+The Stata program imputes CIR the same as J2R, while LMCF differs from all other methods, except that it agrees with MAR at the first visit.
+The R program imputes CIR and J2R differently when in the non-reference arm.
 
@@
 In \texttt{mimix}, they are included in the multivariate normal model fitting as part of the variable $Y_0$.
-[KEVIN IS THIS CORRECT? MIGHT THEY BE WRONGLY CARRIED FORWARD UNDER LMCF?]
 
@@
 & mle 	&	 Logical option to use maximum likelihood parameter estimates instead of MCMC draw parameters. 
-[NEED TO WRITE A COMMENT ABOUT THIS -- KEVIN, OR CAN WE UNDOCUMENT IT?]	\\ \hline
+[WE'RE EXPLORING WHETHER WE CAN HIDE THIS OPTION, WHICH WAS ONLY WRITTEN TO PLEASE JONATHAN BARTLETT]	\\ \hline
 \end{tabular}
@@
 
-The M imputed data sets are output concatenated as one large dataset appended to
-the original unimputed dataset.
-[KEVIN IS IT A DATA SET, A DATA FRAME, OR OTHER?]
+The M imputed data sets are output concatenated as one large data frame, comprising the 
+original unimputed dataset concatenated with all the imptued data sets.
+[CURRENTLY IN WIDE FORMAT, UNLIKE THE INPUT DATA AND UNLIKE STATA'S OUTPUT]
 
@@
 The data are missing after patients discontinue their randomised treatment. 
-A treatment policy estimand is therefore estimated using reference-based imputation.
+A treatment policy estimand is therefore estimated using reference-based imputation, with the placebo arm as reference. 
 For example, J2R is implemented as follows
 \begin{verbatim}
-asthmaJ2R <- mimix(data='asthma', covar=base, depvar=fev, treatvar=treat, idvar=id, timevar=time,
-    method='J2R', reference=1, M=5, seed=101, prior=ridge, burnin=1000)
+asthmaJ2R <- mimix(data=asthma, covar='base', depvar=fev, treatvar=treat, idvar=id, 
+    timevar=time, method='J2R', reference=1, M=5, seed=101, prior=ridge, burnin=1000)
 \end{verbatim}
 
-[KEVIN, IT SEEMS COVAR MUST BE STRING BUT DEPVAR A VARNAME, IS THIS CORRECT? NEEDS A COMMENT?]
-
-Note that the imputed data are in wide format. 
-[KEVIN / ALL - I THINK THIS IS A PROBLEM!?]
 The primary outcome can be analysed using an analysis of covariance by
@@
 \end{verbatim}
-
-[GIVE RESULTS]
+Results are as follows:
+\begin{verbatim}
+         term  estimate  std.error statistic       df      p.value
+1 (Intercept) 0.5821281 0.21097369  2.759245 19.15041 1.242142e-02
+2     treat.L 0.1585223 0.07119548  2.226578 81.64041 2.873017e-02
+3        base 0.6843152 0.09180436  7.454060 28.90698 3.315976e-08
+\end{verbatim}
 
@@
 
-[TO DISCUSS -- TEXT OR TABLE? BOTH ARE INCLUDED BELOW.]
+[TO DISCUSS -- TEXT, TABLE OR BOTH?]
 
@@
 
-Not using these options also means that the LMCF method can be compared with either ALMCF or OLMCF in the Five\_Macros: these are the same in the mimix setting with no covariate interacting with time.
+Not using these options also means that the LMCF method can be compared with either ALMCF or OLMCF in the Five\_Macros.
+[I STILL HAVE TO WORK OUT WHICH WE HAVE IMPLEMENTED.]
 
@@
 Covariates interacting with treatment	&	never	&	always	&	always	\\
-Covariates interacting with time	&	optional	&	never	&	never	\\
+Covariates interacting with time	&	optional	&	always	&	always	\\
 LMCF	&	several flavours: ALMCF, OLMCF, AFCMCF, OFCMCF	&	ALMCF only?	&	ALMCF only?	\\

### 2021-03-13 paper/asthma_example.R

@@ -1,0 +1,5 @@
+asthmaJ2R <- mimix(data=asthma, covar='base', depvar=fev, treatvar=treat, idvar=id, timevar=time,
+                   method='J2R', reference=1, M=5, seed=101, prior=ridge, burnin=1000)
+library(mice)     
+fit <- with(data= as.mids(asthmaJ2R), lm(fev.12 ~ treat + base))
+summary(pool(fit))

### 2021-03-13 test/Ian installation.R

@@ -5,6 +5,8 @@ if(!require(devtools)) install.packages('devtools')
 library(devtools) 
 install_url('https://cran.r-project.org/src/contrib/Archive/norm2/norm2_2.0.3.tar.gz')
 install_github("UCL/mimix")
+# 12mar2021: Kevin has put latest package on developer branch
+install_github("UCL/mimix", ref="developer")
 # couldn't update package: vectrs, mice
 
 # remember that mimix sessions need to include:

### 2021-03-13 test/Ian to compare with Stata.R

@@ -24,23 +24,24 @@ asthma$fev<-asthma$fev*1000
 head(asthma)
 
 # explore data (using skills learned from Michelle)
+library(tidyverse)
 asthma %>% count(treat)
 asthma %>% filter(!is.na(fev)) %>% 
   group_by(treat, time) %>% 
   summarise(n=n(), fevmean=mean(fev), fevsd=sd(fev))
 
 # J2R
-impJ2R1 <- mimix(data="asthma",
+impJ2R1 <- mimix(data=asthma,
   covar=c("base","base2"),
-  depvar="fev",
-  treatvar="treat",
-  idvar="id",
-  timevar="time",
+  depvar=fev,
+  treatvar=treat,
+  idvar=id,
+  timevar=time,
   M=100,
   reference=1,
   method="J2R",
   seed=101,
-  prior="jeffreys",
+  prior=jeffreys,
   burnin=1000,
   bbetween=NULL,
   methodvar=NULL

### 2021-03-13 test/allmissing_R.R

@@ -1,0 +1,73 @@
+# Ian's program to test what happens with no observed data
+# 12mar2021
+
+# Install mimix if required
+if(!require(mimix)) {
+  if(!require(devtools)) install.packages('devtools') 
+  library(devtools) 
+  install_github("UCL/mimix")
+}
+
+# Load mimix
+library(mimix)
+packageVersion("mimix")
+library(mice)
+packageVersion("mice")
+
+setwd("C:/ado/ian/test_Rmimix")
+
+# Open data
+load("C:/ado/ian/Rmimix/data/asthma.RData")
+head(asthma)
+
+# make id=5001 wholly missing
+asthma[1:4,"fev"]<-NA
+
+methods<-c("MAR","J2R","CR","CIR","LMCF")
+imp1 <- list()
+imp2 <- list()
+for(i in 1:length(methods)){
+  print(methods[i])
+  imp1[[i]] <- mimix(data=asthma,
+                     covar="base",
+                     depvar=fev,
+                     treatvar=treat,
+                     idvar=id,
+                     timevar=time,
+                     M=1,
+                     reference=1,
+                     method=methods[i],
+                     seed=101,
+                     prior=jeffreys,
+                     burnin=1000,
+                     bbetween=NULL,
+                     methodvar=NULL
+  )
+  imp2[[i]] <- mimix(data=asthma,
+                     covar="base",
+                     depvar=fev,
+                     treatvar=treat,
+                     idvar=id,
+                     timevar=time,
+                     M=1,
+                     reference=2,
+                     method=methods[i],
+                     seed=101,
+                     prior=jeffreys,
+                     burnin=1000,
+                     bbetween=NULL,
+                     methodvar=NULL
+  )
+}
+print("Ref=1")
+for(i in 1:length(methods)){
+  print(methods[i])
+  print(imp1[[i]] %>% filter(.id==5001))
+}
+print("Ref=2")
+for(i in 1:length(methods)){
+  print(methods[i])
+  print(imp2[[i]] %>% filter(.id==5001))
+}
+# LMCF doesn't impute no-obs reference-arm individuals under MAR
+# All methods don't impute no-obs active-arm individuals under MAR

### 2021-03-22 paper/paperRmimix.tex
New test programs for wholly-missing outcomes; updates to paper after Matteo's comments

@@
 % !TeX root = RJwrapper.tex
+
+% comments from Matteo, 18/3/2021
 
@@
 % software 
-The RBI methods of \citet{Carpenter++13} were implemented in SAS in what have become known as ``the five macros'' and are available on the web page (on www.missingdata.org.uk) of the DIA working group for missing data.
-These macros also provide for ``delta adjustment'' in which imputed values are modified by a user-specified amount \citep{Ratitch++13}: this is useful is performing sensitivity analysis, since the RBI methods make a number of untestable assumptions \citep{ian:ZAMSTAR}.
+The RBI methods of \citet{Carpenter++13} were implemented in SAS in what have become known as ``the five macros'' and are available on the web page (on 
+https://www.lshtm.ac.uk/research/centres-projects-groups/missing-data) [JAMES is this correct?]
+% was www.missingdata.org.uk
+of the DIA working group for missing data.
+These macros also provide for ``delta adjustment'' in which imputed values are modified by a user-specified amount \citep{Ratitch++13}: this is useful in performing sensitivity analysis, since the RBI methods make a number of untestable assumptions \citep{ian:ZAMSTAR}.
 The methods were then implemented in Stata by \citet{Cro++16}.
@@
 % Aims of paper
-We have implemented the RBI methods in a new R package, which includes the a full implementation of the causal model. 
+We have implemented the RBI methods in a new R package, which includes a full implementation of the causal model. 
 The causal model was previously specified for a two-arm trial and has been extended here for a multi-arm trial.
@@
 we assume that quantitative outcome measurements are scheduled at baseline and at $\tmax$ occasions after randomisation. 
-Let $Z$ be the random variable for the participant's randomised treatment arm, where $Z=z$ denotes an arm in which we want to impute missing values, and $Z=r$ denotes the reference arm.
+Let $Z$ be the random variable for the participant's randomised treatment arm, where $Z=z$ denotes any arm (active treatment or reference) in which we want to impute missing values, and $Z=r$ denotes the reference arm.
 Let $Y_t$ be the random variable for the participant's outcome at visit $t=0,...,\tmax$. 
@@
  as before.
-
-[Omit?:] 
-The variance-covariance matrix  of the potential  outcomes is $\boldsymbol{\Sigma}(s)=\varb{\boldsymbol{Y}(s)}$ with submatrices 
-$\boldsymbol{\Sigma}_{\pret\pret}(s)$,
-$\boldsymbol{\Sigma}_{\postt\postt}(s)$ and
-$\boldsymbol{\Sigma}_{\postt\pret}(s)$.
+%The variance-covariance matrix  of the potential  outcomes is $\boldsymbol{\Sigma}(s)=\varb{\boldsymbol{Y}(s)}$ with submatrices 
+%$\boldsymbol{\Sigma}_{\pret\pret}(s)$,
+%$\boldsymbol{\Sigma}_{\postt\postt}(s)$ and
+%$\boldsymbol{\Sigma}_{\postt\pret}(s)$.
 We define 
-the matrix of regression coefficients of  potential outcomes after visit  $t$ on those up to visit  $t$
-as $\boldsymbol{\beta}_t(s) = \boldsymbol{\Sigma}_{\postt\pret}(s) \boldsymbol{\Sigma}_{\pret\pret}(s)^{-1}$, and 
-the residual variance of the potential outcomes after visit  $t$ given those up to visit  $t$ as
-$\boldsymbol{\Omega}_t(s) = \boldsymbol{\Sigma}_{\postt\pret}(s)
-\boldsymbol{\Sigma}_{\pret\pret}(s)^{-1}
-\boldsymbol{\Sigma}_{\postt\pret}(s)^T$.
+the matrix of regression coefficients of  potential outcomes after visit  $t$, $\boldsymbol{Y}_{\postt}(z,s)$, on those up to visit  $t$,  $\boldsymbol{Y}_{\pret}(z,s)$,
+as $\boldsymbol{\beta}_t(z,s)$.
+%  = \boldsymbol{\Sigma}_{\postt\pret}(s) \boldsymbol{\Sigma}_{\pret\pret}(s)^{-1}$, and 
+%the residual variance of the potential outcomes after visit  $t$ given those up to visit  $t$ as
+%$\boldsymbol{\Omega}_t(s) = \boldsymbol{\Sigma}_{\postt\pret}(s)
+%\boldsymbol{\Sigma}_{\pret\pret}(s)^{-1}
+%\boldsymbol{\Sigma}_{\postt\pret}(s)^T$.
 
@@
 \end{eqnarray}
-
-[Again we can handle variances... how much to say?]
+The variance matrix for the imputation model is drawn from the reference arm as described in \citet{ian:RBIcausal}.
 
@@
 
-[TEXT FOR DISCUSSION -- NEED TO WORK OUT HOW TO HANDLE THESE DISCREPANCIES]
+[TEXT FOR DISCUSSION / OMISSION. SUZIE CAN YOU EXPLAIN WHAT THE CODE DOES FOR CIR/LMCF?]
 
-Individuals with no observed values are easily imputed under MAR, but for the other methods they are currently a problem that we need to resolve.
-
-We first consider J2R and CR. 
-These methods should construct the imputation distribution wholly from the reference arm. 
-They therefore correspond to MAR imputation if the wholly unobserved individual was in the reference arm. 
-Both the Stata and R programs do this correctly for an individual in the reference arm, but for an individual in a non-reference arms, J2R and CR do not agree with each other.
-
-We next consider CIR and LMCF. These are defined in terms of the last observed value, so they are formally undefined. 
-The Stata program imputes CIR the same as J2R, while LMCF differs from all other methods, except that it agrees with MAR at the first visit.
-The R program imputes CIR and J2R differently when in the non-reference arm.
+The algorithm described above applies naturally to individuals with no observed values under MAR, J2R and CR. 
+The other methods, CIR, LMCF and the causal model all depend on a last observed value which does not exist. However they all appear to give sensible values.
 
@@
 
-Delta adjustment allows imputations to differ sytematically from RBI methods and so provides a framework for performing sensitivity analysis around RBI assumptions.
+Delta adjustment allows imputations to differ syStematically from RBI methods and so provides a framework for performing sensitivity analysis around RBI assumptions.
 We follow the approach of the five macros. 
@@
 \item Impute post-discontinuation missing values under the user-specified assumption, looping over treatments and patterns.
-If methodvar and/or refvar is specified then the loop is also over teh values of these.
+If methodvar and/or refvar is specified then the loop is also over the values of these.
 \item Perform delta adjustment if specified.
@@
 \end{enumerate}
+
 The baseline value of the outcome could be handled as an outcome or as a covariate. 
-[Which do we recommend?]
+[Which do we recommend? Matteo prefers as a covariate as it's always observed, but if incompete it needs to be an outcome?]
 
@@
 original unimputed dataset concatenated with all the imptued data sets.
-[CURRENTLY IN WIDE FORMAT, UNLIKE THE INPUT DATA AND UNLIKE STATA'S OUTPUT]
 
@@
 \begin{verbatim}
-asthmaJ2R <- mimix(data=asthma, covar='base', depvar=fev, treatvar=treat, idvar=id, 
-    timevar=time, method='J2R', reference=1, M=5, seed=101, prior=ridge, burnin=1000)
+asthmaJ2R <- mimix(data = asthma, covar = 'base', depvar = fev, treatvar = treat, idvar = id, 
+    timevar = time, method = 'J2R', reference = 1, M = 5, seed = 101, prior = ridge, burnin = 1000)
 \end{verbatim}
@@
 library(mice)     
-fit <- with(data= as.mids(asthmaJ2R), lm(fev.12 ~ treat + base))
+fit <- with(data = as.mids(asthmaJ2R), lm(fev.12 ~ treat + base))
 summary(pool(fit))    
@@
 \end{verbatim}
+
+NEEDS MORE EXAMPLE
 
@@
 LMCF	&	several flavours: ALMCF, OLMCF, AFCMCF, OFCMCF	&	ALMCF only?	&	ALMCF only?	\\
-No observed data	&	Warning	&	First mean is used	&	?	\\\hline
+No observed data	&	Warning	&	First mean is used?	&	?	\\\hline
 \end{tabular}
@@
 \item We didn't code the \emph{RBI alternative} approach of \citet{ian:RBIcausal} that instead uses $\boldsymbol{\beta}_t(\tmax)$ and $\boldsymbol{\Omega}_t(\tmax)$ for all RBI methods.
+\item As for any MCMC procedure, an adequate number of iterations is required to achieve convergence and hence suitable imputed values. 
+For the multivariate normal model being fitted here, we believe that 1000 iterations are likely to be ample. 
+We have not implemented diagnostic plots for convergence [KEVIN but does norm2 have any that we could borrow?]
 

### 2021-03-22 test/allmissing_R.R

@@ -59,12 +59,15 @@ for(i in 1:length(methods)){
                      methodvar=NULL
   )
 }
-print("Ref=1")
+library(tidyverse)
+
+cat("\nRef=1\n")
 for(i in 1:length(methods)){
   print(methods[i])
   print(imp1[[i]] %>% filter(.id==5001))
 }
-print("Ref=2")
+
+cat("\nRef=2\n")
 for(i in 1:length(methods)){
   print(methods[i])
   print(imp2[[i]] %>% filter(.id==5001))

### 2021-03-22 test/makelog.R

@@ -1,0 +1,5 @@
+# make log file
+setwd("C:/ado/ian/Rmimix/test")
+sink("allmissing_R.log", append=FALSE, split=TRUE)
+source("allmissing_R.R")
+sink()
